---
title: "Diagnostic Tools for B Estimation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diagnostic Tools for B Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 8
)
```

## Introduction

This vignette demonstrates the diagnostic and visualization tools for assessing the quality of effect size (B) estimation in the GLOW framework. These tools work with both `glow_B_estimate` objects (from `get_B()`) and `glow_B_model` objects (from `train_B_model()`). They help you:

- Evaluate model fit quality
- Detect potential outliers
- Compare different estimation methods
- Identify quality control issues

## Basic Usage

### Estimating B with Full Diagnostics

First, run `get_B()` with `return_full = TRUE` to get a complete result object:

```{r eval=FALSE}
library(GLOWr)

# Example: Estimate B for a set of variants
result <- get_B(
  training_trait = "binary",
  training_MAF = training_MAF,
  training_BETA = training_BETA,
  training_P = training_P,
  training_N = training_N,
  target_trait = "binary",
  target_MAF = target_MAF,
  target_case_prop = 0.1,
  method = "both",        # Run both methods for comparison
  return_full = TRUE,     # Get full result object
  verbose = 2
)
```

### Quick Visual Check

The quickest way to assess quality is with the dashboard plot:

```{r eval=FALSE}
# Display 2x2 dashboard of diagnostic plots
plot(result)
```

This creates a 2x2 layout showing:
- Top-left: Fitted model vs training data
- Top-right: Q-Q plot of residuals
- Bottom-left: Residuals vs fitted values
- Bottom-right: Method comparison (if both methods used) or Cook's distance

### Comprehensive Diagnostics

For a detailed diagnostic report:

```{r eval=FALSE}
# Generate comprehensive diagnostics
# get_B_diagnostics() accepts both glow_B_estimate and glow_B_model objects
diag <- get_B_diagnostics(result, show_plots = TRUE, verbose = 1)

# The function automatically displays:
# - Model fit statistics (R², Adjusted R²)
# - Residual analysis
# - Outlier detection results
# - Method comparison (if applicable)
# - Quality control flags
```

### Diagnostics with `train_B_model()`

You can also run diagnostics on a standalone trained model (without predictions):

```{r eval=FALSE}
# Train a model
b_model <- train_B_model(
  training_trait = "binary",
  training_MAF = training_MAF,
  training_BETA = training_BETA,
  method = "beta"
)

# Plot model-level diagnostics (no target data needed)
plot(b_model)
plot(b_model, type = "fit")
plot(b_model, type = "residuals")

# Full diagnostic report from a trained model
diag <- get_B_diagnostics(b_model, show_plots = TRUE)
```

## Visualization Options

### Plot Types

You can create specific plot types:

```{r eval=FALSE}
# Model fit plot only
plot(result, type = "fit")

# Residual diagnostic plots
plot(result, type = "residuals")

# Method comparison plots (requires both methods)
plot(result, type = "comparison")

# Dashboard (default)
plot(result, type = "all")
```

### Method Comparison Plot

When both methods are used, the comparison plot shows:

```{r eval=FALSE}
plot(result, type = "comparison")
```

This creates a 2x2 layout with:
1. Scatterplot: Beta method vs P-value method
2. Bland-Altman plot: Difference vs average
3. Percent difference vs average
4. Distribution of differences

## Interpreting Diagnostics

### Model Fit Statistics

The diagnostic report provides several measures of model quality:

```{r eval=FALSE}
diag <- get_B_diagnostics(result, show_plots = FALSE, verbose = 0)

# R-squared: Proportion of variance explained
# Diagnostics use short names: $beta and $pvalue (not $beta_method)
diag$model_fit$beta$R2       # Higher is better (>0.5 preferred)

# Adjusted R-squared: Accounts for model complexity
diag$model_fit$beta$adj_R2   # Higher is better

# Model formula
diag$model_fit$beta$formula  # Shows transformation used
```

**Interpretation:**
- **R² > 0.5**: Good fit
- **R² 0.3-0.5**: Moderate fit (may benefit from more data)
- **R² < 0.3**: Poor fit (warning issued)

### Residual Analysis

Check residual behavior:

```{r eval=FALSE}
# Residual statistics
diag$residual_stats$mean       # Should be ~0
diag$residual_stats$sd         # Spread of residuals
diag$residual_stats$normality_p # Shapiro-Wilk test p-value

# Non-normal residuals (p < 0.05) suggest:
# - Transformation may be inappropriate
# - Outliers present
# - Non-linear relationship
```

### Outlier Detection

Identify potentially influential observations:

```{r eval=FALSE}
# Number of outliers detected
diag$outliers$n_outliers

# Proportion of data
diag$outliers$proportion  # Warning if >10%

# Which observations are outliers
diag$outliers$indices

# Cook's distance values
diag$outliers$cooks_d
```

**What to do about outliers:**
1. Investigate whether they are data errors
2. Check if they represent real biological phenomena
3. Consider robustness of results with/without them
4. Document decisions about handling outliers

### Method Comparison

When both methods are used:

```{r eval=FALSE}
# Correlation between methods
diag$method_comparison$correlation  # Should be >0.7

# Root mean squared error
diag$method_comparison$rmse

# Percent differences
diag$method_comparison$mean_percent_diff
diag$method_comparison$max_percent_diff
```

**Interpretation:**
- **High correlation (>0.9)**: Methods agree well
- **Moderate correlation (0.7-0.9)**: Reasonable agreement
- **Low correlation (<0.7)**: Methods disagree (warning issued)

**Note**: Large percent differences can occur even with high correlation because methods operate on different scales. Focus on correlation.

### Quality Control Flags

The diagnostic report automatically identifies potential issues:

```{r eval=FALSE}
# Check for QC issues
if (length(diag$qc_flags) > 0) {
  cat("Quality Control Issues:\n")
  for (flag in diag$qc_flags) {
    cat("  -", flag, "\n")
  }
}
```

Common flags:
- "Low R²": Model explains <30% of variance
- "Non-normal residuals": May indicate model misspecification
- "High proportion of outliers": >10% of data flagged
- "Low correlation between methods": Methods disagree

## Model Comparison

Compare different candidate models:

```{r eval=FALSE}
# Compare all candidate models
model_comp <- compare_B_models(
  result,
  which_method = "beta",
  criterion = "R2",
  plot = TRUE
)

# View results
print(model_comp)
```

**Note**: Current implementation has limited functionality. It shows the selected model but cannot re-fit all 8 candidates without original training data. This is a known limitation documented in the code.

## Complete Example

Here's a complete workflow:

```{r eval=FALSE}
# 1. Load data
als_data <- readRDS("path/to/ALS_data.rds")
training_data <- als_data$data

# 2. Extract required fields
training_MAF <- training_data$MAF
training_BETA <- training_data$BETA
training_P <- training_data$P
training_N <- training_data$N

# Remove any invalid values
valid_idx <- complete.cases(data.frame(training_MAF, training_BETA, training_P, training_N)) &
             training_MAF > 0 & training_MAF < 1 &
             abs(training_BETA) > 1e-10

training_MAF <- training_MAF[valid_idx]
training_BETA <- training_BETA[valid_idx]
training_P <- training_P[valid_idx]
training_N <- training_N[valid_idx]

# 3. Define target variants
target_MAF <- seq(0.01, 0.4, by = 0.05)
target_case_prop <- 0.001  # ALS prevalence

# 4. Estimate B with both methods
result <- get_B(
  training_trait = "binary",
  training_MAF = training_MAF,
  training_BETA = training_BETA,
  training_P = training_P,
  training_N = training_N,
  target_trait = "binary",
  target_MAF = target_MAF,
  target_case_prop = target_case_prop,
  method = "both",
  return_full = TRUE,
  verbose = 2
)

# 5. Visual diagnostics
plot(result)

# 6. Detailed diagnostics
diag <- get_B_diagnostics(result, show_plots = TRUE)

# 7. Check for issues
if (length(diag$qc_flags) > 0) {
  cat("Issues detected:\n")
  print(diag$qc_flags)
}

# 8. Use the estimates
B_estimates <- result$B  # Primary estimates
# Use these in optimal weight calculation...
```

## Recommendations

### For Publication-Quality Analysis

1. **Always use both methods** (when data permits) to validate estimates
2. **Document QC decisions** about outliers and model selection
3. **Report model fit statistics** (R², Adjusted R²) in methods section
4. **Include diagnostic plots** in supplementary materials
5. **Check correlation between methods** - high correlation supports validity

### For Exploratory Analysis

1. **Start with dashboard plot** for quick assessment
2. **Run full diagnostics** to identify issues
3. **Investigate outliers** - are they real or errors?
4. **Compare methods** if both available
5. **Consider sensitivity analysis** with/without outliers

### When Things Go Wrong

If diagnostics reveal issues:

1. **Low R²**:
   - Check if you have enough training data (>50 variants recommended)
   - Consider whether MAF-effect size relationship is expected
   - Try different model selections (though current interface is limited)

2. **Many outliers**:
   - Verify data quality
   - Check for population stratification
   - Consider biological reasons (e.g., large effect variants in specific genes)

3. **Methods disagree**:
   - Check that you're using appropriate trait types
   - Verify SE or case_prop values are correct
   - Consider whether cross-trait transfer is appropriate

4. **Non-normal residuals**:
   - May indicate model misspecification
   - Check for outliers
   - Consider if relationship is truly monotonic

## Summary

The diagnostic tools provide:
- **Visual assessment** through multiple plot types
- **Quantitative metrics** for model quality
- **Automatic QC flagging** of potential issues
- **Method comparison** when multiple approaches used

Use these tools to ensure your B estimates are reliable and appropriate for downstream GLOW analyses.

## See Also

- `?get_B` - All-in-one B estimation (train + predict)
- `?train_B_model` - Train B estimation model separately
- `?predict_B` - Predict B from a trained model
- `?plot.glow_B_estimate` - Plot method for estimation results
- `?plot.glow_B_model` - Plot method for trained models
- `?get_B_diagnostics` - Diagnostic function (accepts both classes)
- `?compare_B_models` - Model comparison documentation
