---
title: "Data Preparation for B Estimation"
author: "GLOWr Package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Preparation for B Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE  # Set to TRUE when package data is available
)
```

## Introduction

The `prepare_B_training_data()` function provides a comprehensive pipeline for preparing GWAS summary statistics for use in B parameter estimation via `get_B()`. This vignette demonstrates how to load, standardize, filter, and validate GWAS data from various sources and formats.

### What This Function Does

The data preparation pipeline handles:

- **Multiple file formats**: Excel (.xlsx, .xls), CSV, TSV, and R data frames
- **Smart column detection**: Automatically recognizes 100+ variations of standard column names
- **Intelligent data conversions**: Extracts numeric values from text, converts OR to BETA, handles -log10(P)
- **Trait type detection**: Sophisticated per-row detection for mixed-trait datasets
- **Flexible filtering**: Filter by study name, PMID, chromosome, or any custom field
- **Quality control**: Sample size thresholds, P-value filters, MAF ranges, duplicate removal
- **Complete validation**: Ensures data meets requirements for B estimation

### Key Features

- **Fully automated**: Minimal user input required for standard GWAS formats
- **Highly customizable**: Override any default behavior with explicit parameters
- **Transparent processing**: Complete metadata tracking of all transformations
- **Publication-ready**: Produces standardized, validated datasets

---

## Quick Start

### Basic Usage

Load GWAS data with minimal parameters:

```{r quickstart-basic}
library(GLOWr)

# Load data from Excel file
gwas_data <- prepare_B_training_data(
  data = "my_gwas_results.xlsx",
  trait_type = "binary",
  verbose = 2
)

# View the results
print(gwas_data)
```

The function returns an S3 object with two components:

- `data`: Cleaned, standardized data frame ready for `get_B()`
- `metadata`: Complete record of processing steps and conversions applied

### Supported Data Formats

```{r formats}
# Excel files (.xlsx, .xls)
data_excel <- prepare_B_training_data(
  data = "gwas_summary.xlsx",
  trait_type = "continuous"
)

# CSV files
data_csv <- prepare_B_training_data(
  data = "gwas_summary.csv",
  trait_type = "binary"
)

# TSV files
data_tsv <- prepare_B_training_data(
  data = "gwas_summary.tsv",
  trait_type = "continuous"
)

# R data frames (already loaded)
my_df <- read.csv("gwas_data.csv")
data_df <- prepare_B_training_data(
  data = my_df,
  trait_type = "binary"
)
```

---

## Column Mapping

### Automatic Column Detection

The function automatically recognizes over 100 variations of standard column names:

| Standard Name | Common Variations |
|---------------|-------------------|
| `rsID` | SNP, SNPS, RSID, RS_ID, VARIANT_ID, ID, MARKERNAME |
| `CHR` | CHR, CHROM, CHROMOSOME, CHR_ID, #CHR |
| `POS` | POS, POSITION, BP, CHR_POS, BASE_PAIR_LOCATION |
| `MAF` | MAF, RAF, FREQ, EAF, AF, RISK.ALLELE.FREQUENCY |
| `P` | P, PVALUE, P_VALUE, P.VALUE, P-VALUE, PVAL |
| `N` | N, SAMPLE_SIZE, SAMPLE.SIZE, N_TOTAL, NEFF |
| `BETA` | BETA, EFFECT, EFFECT_SIZE, B, BETA_NUMERIC |
| `OR` | OR, ODDS_RATIO, ODDSRATIO |

### Custom Column Mapping

Override automatic detection when your column names are non-standard:

```{r column-mapping}
# Specify custom mappings
data <- prepare_B_training_data(
  data = "my_gwas.xlsx",
  trait_type = "continuous",
  column_mapping = list(
    rsID = "SNP_ID",              # Your column: "SNP_ID" -> Standard: "rsID"
    CHR = "CHROMOSOME",           # Your column: "CHROMOSOME" -> Standard: "CHR"
    MAF = "FREQ",                 # Your column: "FREQ" -> Standard: "MAF"
    P = "PVALUE",                 # Your column: "PVALUE" -> Standard: "P"
    N = "SAMPLE_SIZE",            # Your column: "SAMPLE_SIZE" -> Standard: "N"
    BETA = "Effect_Size"          # Your column: "Effect_Size" -> Standard: "BETA"
  ),
  verbose = 2
)
```

**Note**: Unmapped columns are retained in the output with their original names.

---

## Trait Type Detection

### How the trait_type Parameter Works

The function provides sophisticated per-row trait type detection:

- **`trait_type = NULL`** (default): Auto-detect each row based on evidence signals
- **`trait_type = "binary"` or `"continuous"`**:
  - Serves as default for rows with insufficient evidence ("unknown")
  - **High-confidence signals override this default** (e.g., if OR column is present,
    the variant is classified as binary regardless of your trait_type setting)
- **Result**: A TRAIT_TYPE column is added to output with per-variant classifications

**Important**: High-confidence detection signals (like presence of OR values or
Case/Control sizes) take precedence over user-specified trait_type. The user
setting only applies when evidence is weak or absent.

### Overview

The function includes sophisticated **per-row trait type detection** that can handle mixed-trait datasets (e.g., GWAS Catalog data containing both binary and continuous traits). Each variant is classified individually based on available evidence.

### Detection Signals

The function uses multiple evidence sources in order of decreasing confidence:

#### High Confidence Binary Indicators

1. **OR column with value** (excluding ambiguous names like "OR.or.BETA")
   - Presence of odds ratio strongly indicates binary trait

2. **Case Size + Control Size columns both present**
   - Unambiguous indicator of case-control study

3. **Disease name in trait column**
   - Keywords: sclerosis, diabetes, cancer, carcinoma, tumor, disease, disorder, syndrome, osteoporosis, arthritis, alzheimer, parkinson, asthma, schizophrenia, depression, hypertension, coronary, stroke, infection, etc.

4. **"Susceptibility" keyword in annotations**
   - Indicates genetic susceptibility to disease

#### High Confidence Continuous Indicators

1. **Quantitative trait name**
   - Keywords: bone mineral density, BMD, BMI, height, weight, LDL, HDL, cholesterol, triglyceride, glucose, blood pressure, waist, hip ratio, lung function, etc.

2. **BETA present with no binary indicators**
   - Lower confidence; only used when OR column absent or ambiguous

### Conservative Approach

- **Defaults to "unknown"** when evidence is insufficient
- Prioritizes avoiding false classifications over complete classification
- Users can override with explicit `trait_type` parameter

### Usage Examples

#### Auto-detect mixed GWAS Catalog data

```{r trait-type-auto}
# Let function detect trait type per row
result <- prepare_B_training_data(
  data = "gwas_catalog.csv",
  trait_type = NULL,  # Auto-detect per row
  verbose = 2
)

# Check distribution of detected trait types
table(result$data$TRAIT_TYPE)
#   binary continuous unknown
#       45         30       0
```

#### Override with default for unknown cases

```{r trait-type-override}
# Provide default for uncertain cases
result <- prepare_B_training_data(
  data = "my_data.csv",
  trait_type = "binary",  # Used as default for "unknown" cases only
  verbose = 2
)

# High-confidence continuous traits still detected despite default
```

#### ALS data (automatically detects binary)

```{r trait-type-als}
als_result <- prepare_B_training_data(
  data = "ALS-known-SNPs.xlsx",
  trait_type = NULL,
  verbose = 2
)

# Result: All variants correctly detected as binary
# Signals used: Case Size + Control Size, trait name "sclerosis", OR values
```

### Important Notes on Detection

**When Detection is Reliable:**

- Clear binary indicators (OR values, Case+Control, disease names)
- Well-known quantitative traits (BMI, BMD, height)
- Standard GWAS Catalog format

**When to Verify Manually:**

- Custom datasets with non-standard column names
- Ambiguous trait names
- Novel or rare phenotypes not in knowledge base
- Many variants classified as "unknown"

**Detection Limitations:**

- This detection is robust but **NOT 100% accurate**
- Manual verification recommended for critical analyses
- Less critical when using P-value-based (vs. BETA-based) methods for B estimation

---

## Quality Control Filtering

### Default QC Filters

The function applies the following quality control filters by default:

```{r qc-defaults}
# Default QC filters
default_qc <- list(
  min_sample_size = 500,       # Minimum N
  max_pvalue = 1,              # Maximum P (no filter by default)
  min_maf = 0,                 # Minimum MAF (no filter by default)
  remove_duplicates = TRUE,    # Remove duplicate variants
  duplicate_priority = "N",    # Keep variant with largest N
  remove_na = TRUE,            # Remove rows with NA
  remove_na_cols = c("MAF", "P", "N")  # Columns that must be non-NA
)
```

### Customizing QC Filters

#### Stricter QC

```{r qc-strict}
# More stringent filtering
data <- prepare_B_training_data(
  data = "gwas_data.xlsx",
  trait_type = "continuous",
  qc_filters = list(
    min_sample_size = 5000,      # Require larger studies
    max_pvalue = 5e-8,           # Genome-wide significance only
    min_maf = 0.01,              # Remove rare variants
    remove_duplicates = TRUE,
    remove_na = TRUE,
    remove_na_cols = c("MAF", "P", "N", "BETA")  # Also require BETA
  ),
  verbose = 2
)
```

#### More Lenient QC

```{r qc-lenient}
# Minimal filtering
data <- prepare_B_training_data(
  data = "gwas_data.xlsx",
  trait_type = "binary",
  qc_filters = list(
    min_sample_size = 100,       # Accept smaller studies
    max_pvalue = 1,              # No P-value filter
    min_maf = 0,                 # Include all MAFs
    remove_duplicates = TRUE,
    remove_na = TRUE,
    remove_na_cols = c("MAF", "P", "N")  # Only basic required columns
  ),
  verbose = 2
)
```

### Understanding NA Filtering

The `remove_na_cols` parameter specifies which columns must have non-NA values:

```{r qc-na}
# Example: Require both BETA and P to be non-NA
data <- prepare_B_training_data(
  data = "gwas_data.xlsx",
  trait_type = "binary",
  qc_filters = list(
    remove_na = TRUE,
    remove_na_cols = c("MAF", "P", "N", "BETA")  # All four must be non-NA
  ),
  verbose = 2
)
```

### Duplicate Removal

Duplicates are removed by **both rsID and genomic position (CHR:POS)**:

```{r qc-duplicates}
# Duplicates removed, keeping highest N
data <- prepare_B_training_data(
  data = "gwas_with_duplicates.xlsx",
  trait_type = "continuous",
  qc_filters = list(
    remove_duplicates = TRUE,
    duplicate_priority = "N"  # Keep variant with largest sample size
  ),
  verbose = 2
)
```

**How it works:**

1. Removes duplicate rsIDs (keeps highest N)
2. Removes duplicate CHR:POS positions (keeps highest N)
3. Reports number removed in each category

---

## General Field-Based Filtering

### Overview

The flexible filtering system allows you to filter on **any column** in your data, not just standard GWAS columns. This is useful for:

- Selecting specific studies or publications
- Excluding overlapping samples
- Filtering by ancestry, trait, annotation, or custom fields

### Inclusion Filters

Keep only rows matching specified values:

```{r filter-include-studies}
# Include only specific studies
bmd_data <- prepare_B_training_data(
  data = "bmd_gwas.xlsx",
  trait_type = "continuous",
  filter_include = list(
    STUDY = c("Atlas", "153 new loci", "613 new loci")
  ),
  verbose = 2
)
```

```{r filter-include-chr}
# Include only autosomes
data <- prepare_B_training_data(
  data = "gwas_data.csv",
  trait_type = "binary",
  filter_include = list(
    CHR = 1:22  # Chromosomes 1-22 only
  ),
  verbose = 2
)
```

### Exclusion Filters

Remove rows matching specified values:

```{r filter-exclude-pmid}
# Exclude specific publications (to avoid overlapping samples)
data <- prepare_B_training_data(
  data = "my_gwas.csv",
  trait_type = "binary",
  filter_exclude = list(
    PMID = c("12345678", "87654321")  # Exclude these PMIDs
  ),
  verbose = 2
)
```

```{r filter-exclude-trait}
# Exclude specific trait annotations
als_data <- prepare_B_training_data(
  data = "ALS-known-SNPs.xlsx",
  trait_type = "binary",
  filter_exclude = list(
    `Reported trait` = c(
      "Amyotrophic lateral sclerosis (C9orf72 mutation interaction)",
      "Amyotrophic lateral sclerosis (age of onset)",
      "Survival in sporadic amyotrophic lateral sclerosis"
    ),
    `P-value annotation` = c("(age of onset)", "(survival)", "(site of onset)")
  ),
  verbose = 2
)
```

### Pattern Matching vs. Exact Matching

```{r filter-pattern}
# Pattern matching (default)
data <- prepare_B_training_data(
  data = "gwas_catalog.csv",
  trait_type = "binary",
  filter_include = list(
    STUDY = "diabetes"  # Matches "Type 2 diabetes", "Diabetes mellitus", etc.
  ),
  filter_pattern = TRUE,  # Use grepl for character columns
  verbose = 2
)

# Exact matching
data <- prepare_B_training_data(
  data = "gwas_catalog.csv",
  trait_type = "binary",
  filter_include = list(
    STUDY = "Type 2 diabetes"  # Exact match only
  ),
  filter_pattern = FALSE,  # Use %in% for exact matching
  verbose = 2
)
```

### Combined Filtering

Combine inclusion and exclusion filters:

```{r filter-combined}
# Complex filtering scenario
data <- prepare_B_training_data(
  data = "gwas_catalog.csv",
  trait_type = "binary",
  filter_include = list(
    CHR = 1:22,                 # Autosomes only
    STUDY = c("Study1", "Study2", "Study3")  # Specific studies
  ),
  filter_exclude = list(
    PMID = c("11111111", "22222222"),  # Exclude overlapping publications
    Population = "Asian"                # Exclude specific ancestry
  ),
  filter_pattern = TRUE,
  verbose = 2
)
```

**Note**: Inclusion filters take precedence over exclusion filters for the same column.

---

## Data Conversions

### Overview

The function performs intelligent data conversions automatically to handle real-world GWAS data formats:

1. Extract sample sizes from text descriptions
2. Extract P-values from text (scientific notation)
3. Convert -log10(P) to P-values
4. Convert OR to log(OR) for binary traits
5. Harmonize MAF values (MAF > 0.5 becomes 1 - MAF)
6. Convert CHR to numeric (X→23, Y→24, MT→25, then filter to autosomes only)

### Sample Size Extraction from Text

```{r conversion-sample-size}
# Handles text-based sample sizes
data <- prepare_B_training_data(
  data = "gwas_with_text.xlsx",
  trait_type = "continuous",
  extract_text_values = TRUE,  # Default
  verbose = 2
)

# Conversions applied:
# "66,420 European ancestry men" -> 66420
# "5,000 cases + 5,000 controls" -> 10000
# "2,678 Icelandic cases" -> 2678
```

### P-value Extraction from Text

```{r conversion-pvalue}
# Handles various P-value formats
data <- prepare_B_training_data(
  data = "gwas_with_special_chars.xlsx",
  trait_type = "binary",
  extract_text_values = TRUE,
  verbose = 2
)

# Conversions applied:
# "5x10^-8" -> 5e-8
# "5E-8" -> 5e-8
# "< 0.001" -> 0.001 (conservative)
```

### -log10(P) to P-value Conversion

```{r conversion-mlog-p}
# Automatically detects and converts PVALUE_MLOG column
data <- prepare_B_training_data(
  data = "gwas_with_mlog_p.xlsx",
  trait_type = "continuous",
  verbose = 2
)

# Conversion: P = 10^(-PVALUE_MLOG)
# PVALUE_MLOG = 8 -> P = 1e-8
```

### OR to BETA Conversion

```{r conversion-or}
# Automatically converts OR to log(OR) for binary traits
data <- prepare_B_training_data(
  data = "case_control_study.xlsx",
  trait_type = "binary",
  convert_or_to_beta = TRUE,  # Default for binary traits
  verbose = 2
)

# Conversion: BETA = log(OR)
# OR = 1.5 -> BETA = log(1.5) = 0.405
```

### MAF Harmonization

```{r conversion-maf}
# MAF values > 0.5 automatically converted to 1 - MAF
data <- prepare_B_training_data(
  data = "gwas_data.xlsx",
  trait_type = "continuous",
  verbose = 2
)

# Example: MAF = 0.8 -> MAF = 0.2
```

**Performed automatically in Step 6 of the pipeline (before QC filtering):**

MAF values > 0.5 are automatically converted to 1 - MAF to ensure all
frequency values represent the minor (less common) allele. This happens
before any QC filters are applied, ensuring your `min_maf` threshold works
correctly.

### Chromosome Encoding

**Automatic Processing:**

The function automatically handles various chromosome encodings:

1. Recognizes sex chromosomes: "X", "chrX", "chr23" → identified as X
2. Recognizes Y chromosomes: "Y", "chrY", "chr24" → identified as Y
3. Recognizes mitochondrial: "MT", "M", "chr25", "chrMT" → identified as MT
4. **Removes all non-autosomal variants (keeps only CHR 1-22)**
5. **Final dataset contains ONLY autosomes (chromosomes 1-22)**

This filtering is automatic and cannot be disabled.

```{r conversion-chr}
# Example: Chromosome encoding is handled automatically
data <- prepare_B_training_data(
  data = "gwas_with_chr_names.xlsx",
  trait_type = "binary",
  verbose = 2
)

# The function automatically:
# - Recognizes and converts various chromosome formats
# - Removes all non-autosomal chromosomes (X, Y, MT)
# - Retains only chromosomes 1-22
```

### Disabling Conversions

```{r conversion-disable}
# Disable automatic conversions if needed
data <- prepare_B_training_data(
  data = "gwas_data.xlsx",
  trait_type = "binary",
  convert_or_to_beta = FALSE,     # Don't convert OR to BETA
  extract_text_values = FALSE,    # Don't extract from text
  verbose = 2
)
```

---

## Advanced Usage and Examples

### Example 1: ALS GWAS Data (Binary Trait)

**Note**: The example files referenced below are not included in the package.
Replace file paths with your own GWAS data files. The code structure and
parameters shown are realistic and can be adapted to your data.

Complete workflow for ALS data with BETA-based B estimation:

```{r example-als, eval=FALSE}
library(GLOWr)

# Load and prepare ALS data
als_data <- prepare_B_training_data(
  data = "ALS-known-SNPs.xlsx",
  trait_type = "binary",
  column_mapping = list(
    BETA = "Beta_numeric"  # File has both "Beta" and "Beta_numeric"
  ),
  filter_exclude = list(
    `Reported trait` = c(
      "Amyotrophic lateral sclerosis (C9orf72 mutation interaction)",
      "Amyotrophic lateral sclerosis (age of onset)",
      "Survival in sporadic amyotrophic lateral sclerosis"
    ),
    `P-value annotation` = c("(age of onset)", "(survival)", "(site of onset)")
  ),
  qc_filters = list(
    min_sample_size = 500,
    remove_duplicates = TRUE,
    remove_na = TRUE,
    remove_na_cols = c("MAF", "P", "N", "BETA")  # BETA required for ALS
  ),
  verbose = 2
)

# Check the results
print(als_data)
# GLOW Training Data for B Estimation
# ====================================
# Metadata:
#   Trait type:        binary
#   Data format:       excel
#   Variants:         143 (209 removed during QC)
#   Date prepared:     2025-11-16

# View data
head(als_data$data[, 1:8])
#                      rsID CHR       POS MAF     P    N  BETA TRAIT_TYPE
# 1              rs2364403   1 155957961 0.2 1e-06 4243    NA     binary
# 2  rs3177980 (rs1131498)   1 169707345 0.3 4e-06 4079 0.509     binary
# 3              rs1541160   1 170026661 0.3 2e-08 4079 0.580     binary

# Use for B estimation with BETA method
B_estimates <- get_B(
  training_trait = "binary",
  training_MAF = als_data$data$MAF,
  training_BETA = als_data$data$BETA[!is.na(als_data$data$BETA)],
  training_P = als_data$data$P,
  training_N = als_data$data$N,
  target_trait = "binary",
  target_MAF = seq(0.01, 0.4, by = 0.05),
  target_case_prop = 0.258
)
```

### Example 2: BMD GWAS Data (Continuous Trait)

**Note**: Replace example file path with your own data file. The filtering
approach shown (inclusion by study keywords) works with any dataset containing
a study identifier column.

Complete workflow for BMD data with P-value-based B estimation:

```{r example-bmd, eval=FALSE}
library(GLOWr)

# Load and prepare BMD data
bmd_data <- prepare_B_training_data(
  data = "literature_snps_filtered.xlsx",
  trait_type = "continuous",
  filter_include = list(
    STUDY = c("Atlas", "153 new loci", "613 new loci")
  ),
  qc_filters = list(
    min_sample_size = 1000,
    remove_duplicates = TRUE,
    remove_na = TRUE,
    remove_na_cols = c("MAF", "P", "N")  # BETA not required for P-value method
  ),
  extract_text_values = TRUE,  # Extract N from text
  verbose = 2
)

# Check the results
print(bmd_data)
# GLOW Training Data for B Estimation
# ====================================
# Metadata:
#   Trait type:        continuous
#   Data format:       excel
#   Variants:         599 (2071 removed during QC)
#   Date prepared:     2025-11-16

# Check metadata for conversions applied
bmd_data$metadata$conversions_applied
# $text_to_N
# [1] TRUE
#
# $OR_to_BETA
# [1] FALSE

# Use for B estimation with P-value method (no BETA needed)
B_estimates <- get_B(
  training_trait = "continuous",
  training_MAF = bmd_data$data$MAF,
  training_P = bmd_data$data$P,
  training_N = bmd_data$data$N,
  target_trait = "continuous",
  target_MAF = seq(0.01, 0.4, by = 0.05),
  target_SE = 0.15
)
```

### Example 3: Custom Workflow with Manual Filtering

Multi-step workflow with custom filtering and inspection:

```{r example-custom, eval=FALSE}
library(GLOWr)

# Step 1: Load data with lenient QC to examine
gwas_data_all <- prepare_B_training_data(
  data = "my_gwas.xlsx",
  trait_type = "binary",
  qc_filters = list(
    min_sample_size = 50,     # Very lenient
    max_pvalue = 1,
    min_maf = 0,
    remove_duplicates = TRUE
  ),
  verbose = 2
)

# Step 2: Examine data distributions
summary(gwas_data_all$data$N)
summary(gwas_data_all$data$P)
summary(gwas_data_all$data$MAF)

# Step 3: Apply stricter QC based on inspection
gwas_data_clean <- prepare_B_training_data(
  data = "my_gwas.xlsx",
  trait_type = "binary",
  qc_filters = list(
    min_sample_size = 2000,    # Decided based on data
    max_pvalue = 5e-8,         # Genome-wide significant
    min_maf = 0.01,            # Common variants
    remove_duplicates = TRUE
  ),
  exclude_studies = c("Pilot Study"),
  verbose = 2
)

# Step 4: Examine QC summary
gwas_data_clean$metadata$qc_summary
# Shows exactly how many variants removed at each step

# Step 5: Further manual filtering if needed
clean_data <- gwas_data_clean$data[
  gwas_data_clean$data$MAF > 0.05 &     # Even more common
  gwas_data_clean$data$P < 1e-10,       # Even more significant
]

# Step 6: Use for B estimation
B_estimates <- get_B(
  training_trait = "binary",
  training_MAF = clean_data$MAF,
  training_BETA = clean_data$BETA,
  training_P = clean_data$P,
  training_N = clean_data$N,
  target_trait = "binary",
  target_MAF = seq(0.01, 0.5, by = 0.01),
  target_case_prop = 0.5
)
```

---

## Two Approaches to Data Preparation

The GLOWr package provides **two valid approaches** for preparing GWAS data for B estimation. Both approaches produce identical results but serve different use cases.

### Approach 1: Universal Function (Recommended for Most Users)

The `prepare_B_training_data()` function provides a **general-purpose, parameter-driven approach** that works with any GWAS dataset format. This is the recommended approach for most users.

**Advantages:**
- One function handles all datasets
- Highly flexible parameter system
- Easy to adapt to new data formats
- Well-documented and maintained
- Comprehensive QC and validation

**Example:**

```{r approach-universal, eval=FALSE}
library(GLOWr)

# Process ALS data using universal function
als_result <- prepare_B_training_data(
  data = "ALS-known-SNPs.xlsx",
  trait_type = "binary",
  column_mapping = list(
    BETA = "Beta_numeric",
    MAF = "RAF",
    P = "P-value",
    N = "Sample Size",
    rsID = "rsID",
    CHR = "chr",
    POS = "pos_GRCh37"
  ),
  filter_exclude = list(
    `Reported trait` = c(
      "Amyotrophic lateral sclerosis (C9orf72 mutation interaction)",
      "Amyotrophic lateral sclerosis (age of onset)"
    ),
    `P-value annotation` = c("(age of onset)", "(survival)")
  ),
  qc_filters = list(
    min_sample_size = 500,
    remove_duplicates = TRUE,
    duplicate_priority = "N",
    remove_na_cols = c("MAF", "P", "N", "BETA")
  ),
  verbose = 2
)

# Access cleaned data
als_data <- als_result$data
```

### Approach 2: Custom Processing Functions

For users who need **maximum control** or have **complex dataset-specific logic**, the package also demonstrates how to write custom processing functions.

**Advantages:**
- Complete manual control over every step
- Can implement complex custom logic
- Good for learning the data preparation process
- Useful when universal function can't meet specific requirements

**Example:**

The package includes example custom functions in `inst/extdata/get_B_example_data.R`:
- `prepare_als_example_data()` - Custom ALS processing
- `prepare_bmd_example_data()` - Custom BMD processing

```{r approach-custom, eval=FALSE}
library(GLOWr)

# Process ALS data using custom function
als_result <- prepare_als_example_data()

# Access cleaned data
als_data <- als_result$data
```

**Note**: These custom functions do **NOT** use `prepare_B_training_data()` internally. They demonstrate independent implementations with dataset-specific logic.

### Comparison

| Aspect | Custom Processing | Universal Function |
|--------|-------------------|-------------------|
| **Functions** | `prepare_als_example_data()`<br>`prepare_bmd_example_data()` | `prepare_B_training_data()` |
| **Flexibility** | Dataset-specific | General-purpose |
| **Control** | Full manual control | Parameter-driven |
| **Code Complexity** | More code, explicit steps | Concise, parameterized |
| **Adaptability** | Requires rewriting for new data | Easy to adapt with new parameters |
| **Use Case** | Complex custom logic | Standard GWAS processing |
| **Output** | Identical processed data | Identical processed data |

### Demonstration: Both Approaches Produce Equivalent Results 

The package includes example scripts demonstrating both approaches:

```{r comparison-demo, eval=FALSE}
# View the custom approach script
file.show(system.file("extdata", "prepare_example_data.R", package = "GLOWr"))

# View the universal approach script (recommended)
file.show(system.file("extdata", "prepare_example_data_universal.R", package = "GLOWr"))
```

Both scripts process the same raw data files and produce identical outputs with the same SNVs, QC filters, and final datasets.

### Which Approach Should You Use?

**Use the Universal Function (`prepare_B_training_data()`) when:**
- Processing standard GWAS data formats
- You want a simple, well-tested approach
- Your filtering needs can be met with parameters
- You're new to the package

**Write Custom Functions when:**
- You need complex, dataset-specific processing logic
- The universal function can't meet your specific requirements
- You want to learn the detailed data preparation process
- You have unique data formats requiring special handling

**Recommendation**: Start with the universal approach (`prepare_B_training_data()`). Only write custom processing functions if you have specific requirements that cannot be met with the flexible parameter system.

---

## Troubleshooting and FAQ

### Problem 1: "Column X not found"

**Cause**: Column name not recognized by automatic detection.

**Solution**:

```{r trouble-column}
# Option 1: Check what columns you actually have
df <- readxl::read_excel("your_file.xlsx")
names(df)

# Option 2: Specify mapping explicitly
data <- prepare_B_training_data(
  data = "your_file.xlsx",
  trait_type = "binary",
  column_mapping = list(
    BETA = "Your_Beta_Column_Name",
    N = "Your_Sample_Size_Column"
  ),
  verbose = 2
)
```

### Problem 2: "All variants removed during QC"

**Cause**: QC filters too strict or data quality issues.

**Solution**:

```{r trouble-qc}
# Start with minimal QC to see what you have
data <- prepare_B_training_data(
  data = "your_file.xlsx",
  trait_type = "binary",
  qc_filters = list(
    min_sample_size = 50,   # Very lenient
    max_pvalue = 1,
    min_maf = 0,
    remove_duplicates = FALSE,
    remove_na = TRUE
  ),
  verbose = 2  # See detailed messages
)

# Check distributions
summary(data$data)

# Then gradually increase stringency based on what you see
```

### Problem 3: "Sample sizes all NA"

**Cause**: Text-based sample sizes not extracted.

**Solution**:

```{r trouble-sample-size}
# Ensure text extraction is enabled (it is by default)
data <- prepare_B_training_data(
  data = "your_file.xlsx",
  trait_type = "continuous",
  extract_text_values = TRUE,
  verbose = 2
)

# If still failing, inspect the N column manually
df <- readxl::read_excel("your_file.xlsx")
head(df$`Sample size`, 20)  # Check format

# You may need to extract manually first
df$N <- extract_sample_size_from_text(df$`Sample size`)

# Then use the modified data frame
data <- prepare_B_training_data(
  data = df,
  trait_type = "continuous"
)
```

### Problem 4: "P-value validation failed"

**Cause**: Invalid P-values in data (P <= 0 or P > 1).

**Solution**:

```{r trouble-pvalue}
# The function automatically removes invalid P-values
data <- prepare_B_training_data(
  data = "your_file.xlsx",
  trait_type = "binary",
  verbose = 2  # Will report how many invalid P-values removed
)

# If you want to inspect first
df <- readxl::read_excel("your_file.xlsx")
table(df$P <= 0 | df$P > 1, useNA = "always")
```

### Problem 5: Multiple columns with similar names

**Cause**: File has both "Beta" (character) and "Beta_numeric" (numeric).

**Solution**:

```{r trouble-multiple}
# Specify which one to use
data <- prepare_B_training_data(
  data = "your_file.xlsx",
  trait_type = "binary",
  column_mapping = list(
    BETA = "Beta_numeric"  # Use the numeric column explicitly
  ),
  verbose = 2
)
```

### Problem 6: Trait type detection incorrect

**Cause**: Ambiguous data or limited detection signals.

**Solution**:

```{r trouble-trait-type}
# Option 1: Explicitly specify trait type
data <- prepare_B_training_data(
  data = "your_file.xlsx",
  trait_type = "binary",  # Override auto-detection
  verbose = 2
)

# Option 2: Check what was detected
result <- prepare_B_training_data(
  data = "your_file.xlsx",
  trait_type = NULL,  # Auto-detect
  verbose = 2
)

# Examine detected types
table(result$data$TRAIT_TYPE)

# Manually fix if needed
result$data$TRAIT_TYPE[result$data$TRAIT_TYPE == "unknown"] <- "binary"
```

---

## Tips and Best Practices

### Tip 1: Always Use `verbose = 2` Initially

See detailed messages about processing steps:

```{r tip-verbose}
data <- prepare_B_training_data(
  data = "your_file.xlsx",
  trait_type = "binary",
  verbose = 2  # Shows all processing details
)
```

### Tip 2: Check Metadata After Processing

Examine what transformations were applied:

```{r tip-metadata}
# View all metadata
names(data$metadata)

# Check conversions
data$metadata$conversions_applied

# Check QC summary
data$metadata$qc_summary

# Check filters applied
data$metadata$filters_applied
```

### Tip 3: Start Lenient, Then Tighten QC

```{r tip-iterative}
# First pass: See what you have
data_all <- prepare_B_training_data(
  data = "your_file.xlsx",
  trait_type = "binary",
  qc_filters = list(
    min_sample_size = 50,
    max_pvalue = 1,
    min_maf = 0
  )
)

# Check distributions
summary(data_all$data$N)
summary(data_all$data$P)
summary(data_all$data$MAF)

# Second pass: Apply appropriate filters based on inspection
data_clean <- prepare_B_training_data(
  data = "your_file.xlsx",
  trait_type = "binary",
  qc_filters = list(
    min_sample_size = 500,    # Decided based on data
    max_pvalue = 0.05,
    min_maf = 0.01
  )
)
```

### Tip 4: Save Processed Data

Avoid re-processing every time:

```{r tip-save}
# Save processed data
saveRDS(gwas_data, "my_gwas_processed.rds")

# Load later
gwas_data <- readRDS("my_gwas_processed.rds")

# Use directly with get_B()
B_estimates <- get_B(
  training_trait = gwas_data$metadata$trait_type,
  training_MAF = gwas_data$data$MAF,
  training_P = gwas_data$data$P,
  training_N = gwas_data$data$N,
  target_trait = "binary",
  target_MAF = seq(0.01, 0.5, by = 0.01)
)
```

### Tip 5: Validate Your Data

Check that data meets requirements:

```{r tip-validate}
# Manual validation check
validation <- validate_B_training_data(
  data = gwas_data$data,
  method = "pvalue",
  verbose = 2
)

if (!validation$valid) {
  print(validation$errors)
}
```

---

## Summary

The `prepare_B_training_data()` function provides a comprehensive, intelligent pipeline for processing GWAS summary statistics:

- **Automatic**: Smart column detection, text extraction, conversions
- **Flexible**: Support for Excel, CSV, TSV, data frames
- **Robust**: Comprehensive QC and validation
- **Transparent**: Full metadata tracking
- **Ready**: Standardized output for `get_B()` estimation

**Key Outputs:**

1. **data**: Clean data frame with standardized columns (rsID, CHR, POS, MAF, P, N, BETA, TRAIT_TYPE, ...)
2. **metadata**: Complete processing record (conversions, filters, QC summary)

**Next Steps:**

After preparing your data, use `get_B()` for B parameter estimation:

```{r summary-workflow, eval=FALSE}
# 1. Prepare data
gwas_data <- prepare_B_training_data(
  data = "my_gwas.xlsx",
  trait_type = "binary",
  verbose = 2
)

# 2. Estimate B parameters
B_estimates <- get_B(
  training_trait = "binary",
  training_MAF = gwas_data$data$MAF,
  training_P = gwas_data$data$P,
  training_N = gwas_data$data$N,
  target_trait = "binary",
  target_MAF = seq(0.01, 0.5, by = 0.01)
)

# 3. Use B estimates in GLOW tests
# See vignette("estimating_effect_sizes") for details
```

---

## Getting Help

- **Function documentation**: `?prepare_B_training_data`
- **Related functions**: `?standardize_column_names`, `?apply_qc_filters`, `?validate_B_training_data`
- **Package vignettes**: `browseVignettes("GLOWr")`
- **B estimation vignette**: `vignette("estimating_effect_sizes", package = "GLOWr")`

---

**Date**: `r Sys.Date()`
**Package**: GLOWr
**Version**: `r packageVersion("GLOWr")`
