---
title: "Estimating Allelic Effect Sizes (B) for Optimal Weighting"
author: "GLOWr Package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating Allelic Effect Sizes (B) for Optimal Weighting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

In the GLOW (inteGrative anaLysis using Optimized Weights) framework, accurate estimation of allelic effect sizes $b(q)$ as a function of minor allele frequency (MAF) $q$ is crucial for calculating optimal weights. This vignette demonstrates how to estimate effect sizes from literature data using both the all-in-one `get_B()` function and the train/predict workflow (`train_B_model()` + `predict_B()`).

### Why Estimate B?

The GLOW framework assigns optimal weights to variants in a gene set based on two key components:

1. **Effect size (B)**: The expected magnitude of the allelic effect given the variant's MAF
2. **Variant-importance score (PI)**: The annotation-derived relative likelihood that a variant is important for the trait

Accurate B estimation improves power by:

- Upweighting variants with larger expected effects
- Appropriately accounting for the MAF-effect size relationship
- Leveraging information from large-scale GWAS studies

### Two Workflows

GLOWr provides two workflows for B estimation:

1. **All-in-one**: `get_B()` trains a model and predicts B in a single call. Best for simple, one-shot analyses.
2. **Train/predict**: `train_B_model()` trains a model once, then `predict_B()` applies it to new MAF values. Best for per-chromosome or repeated-prediction workflows.

### Two Estimation Methods

`get_B()` supports two complementary methods:

**1. Direct Beta Method**

- Uses effect sizes (BETA) directly from training data
- Best when training and target have the same trait type
- Provides the most direct relationship between MAF and effect size
- Formula: Model $\beta^2 \sim f(q)$ where $\beta$ is the effect size and $q$ is MAF

**2. P-value/Z-score Method**

- Uses p-values to compute a trait-independent measure $h(q)$
- Can transfer information between different trait types
- More flexible but requires careful specification of target parameters
- Formula: $Z^2 = \chi^2_{1-p}(1)$, then $h^2(q) = Z^2 / [2Nq(1-q)]$

When both methods are applicable (same trait type, both BETA and p-values available), running both provides validation of estimates.

### What You'll Learn

This vignette covers:

- Quick start with minimal code (both workflows)
- Understanding when to use each method
- Working with real example datasets (ALS, BMD)
- Comparing methods with `method="both"`
- The train/predict workflow for per-chromosome analysis
- Diagnostics and quality control
- Practical recommendations for your analysis

## Quick Start

### All-in-One with `get_B()`

Here's a complete example in ~10 lines:

```{r quickstart, eval=FALSE}
library(GLOWr)

# Load example ALS data (lazy-loaded package data)
data(ALS_snvs_B_training)

# Define target MAFs (your study variants)
target_mafs <- seq(0.01, 0.4, length.out = 50)

# Estimate effect sizes
B <- get_B(
  training_trait = "binary",
  training_MAF = ALS_snvs_B_training$MAF,
  training_BETA = ALS_snvs_B_training$BETA,
  target_trait = "binary",
  target_MAF = target_mafs,
  target_case_prop = 0.001  # ALS case proportion in study
)

# Plot results
plot(target_mafs, B, type = "l", lwd = 2, col = "darkblue",
     xlab = "Minor Allele Frequency", ylab = "Estimated Effect Size (B)",
     main = "Effect Size vs MAF for ALS")
```

That's it! You now have effect size estimates for each target variant that can be used in optimal weight calculation.

### Train Once, Predict Many with `train_B_model()` + `predict_B()`

When you need to predict B for multiple chromosome or variant sets, train the model once and reuse it:

```{r quickstart-train-predict, eval=FALSE}
library(GLOWr)

# Load example ALS data (lazy-loaded package data)
data(ALS_snvs_B_training)

# Step 1: Train the model (once)
b_model <- train_B_model(
  training_trait = "binary",
  training_MAF = ALS_snvs_B_training$MAF,
  training_BETA = ALS_snvs_B_training$BETA,
  method = "beta"
)
print(b_model)

# Step 2: Predict B for different target MAF sets (many times)
# E.g., per chromosome in a genome-wide analysis
for (chr in 1:22) {
  chr_mafs <- get_chromosome_mafs(chr)  # your function to get MAFs
  B_chr <- predict_B(b_model, target_MAF = chr_mafs)
  # ... use B_chr in downstream analysis
}

# The trained model can also be saved/loaded for later use
save_B_model(b_model, "my_trained_B_model.rds")
loaded_model <- load_B_model("my_trained_B_model.rds")
```

The train/predict workflow avoids redundant model fitting and is the recommended approach for pipelines that process data chromosome-by-chromosome.

## Understanding the Methods

### Direct Beta Method: Using Effect Sizes Directly

When you have effect sizes (BETA values) from studies of the same trait, the direct beta method models the relationship:

$$\beta^2 = f(q)$$

where $\beta$ is the effect size and $q$ is the MAF.

**Mathematical Details:**

The function:

1. Squares the training BETA values: $Y = \beta^2$
2. Fits regression models with various transformations
3. Selects the best-fitting model (by default, highest $R^2$)
4. Predicts $\hat{Y}$ for target MAF
5. Returns $\hat{\beta} = \sqrt{\hat{Y}}$

**Eight Candidate Models:**

The function automatically tests 8 transformations:

| Model | Formula | When It Works Best |
|-------|---------|-------------------|
| 1 | $\beta^2 \sim q$ | Linear MAF-effect relationship |
| 2 | $\beta^2 \sim q(1-q)$ | Effect peaks at intermediate MAF |
| 3 | $\beta^2 \sim \log(q)$ | Effect increases with rarer variants |
| 4 | $\beta^2 \sim \log[q(1-q)]$ | Combines frequency function with log |
| 5 | $\log(\beta^2) \sim q$ | Exponential relationship |
| 6 | $\log(\beta^2) \sim q(1-q)$ | Log-linear with frequency function |
| 7 | $\log(\beta^2) \sim \log(q)$ | Power law relationship |
| 8 | $\log(\beta^2) \sim \log[q(1-q)]$ | Complex power law |

**When to Use:**

- Training and target traits are the same type (both binary or both continuous)
- Effect sizes (BETA) are available in training data
- You want the most direct MAF-effect size relationship
- Training data is from well-powered studies

### P-value/Z-score Method: Trait-Independent Estimation

This method converts p-values to a trait-independent measure that can transfer across trait types.

**Mathematical Details:**

The transformation proceeds in steps:

1. **Convert p-values to $Z^2$:**
   $$Z^2 = \chi^2_{1-p}(1) = \text{qchisq}(p, \text{df}=1, \text{lower.tail}=\text{FALSE})$$

2. **Calculate trait-independent $h^2$:**
   $$h^2(q) = \frac{Z^2}{2Nq(1-q)}$$
   where $N$ is the sample size.

3. **Model $h^2$ as function of MAF:**
   $$h^2 = f(q)$$

4. **Convert back to trait-specific $\beta$:**
   - **Continuous trait:** $\beta = h \times \text{SE}^*$
   - **Binary trait:** $\beta = h / \sqrt{p_0^*(1-p_0^*)}$

   where $\text{SE}^*$ is the standard error for continuous traits and $p_0^*$ is the case proportion in the target case-control study.

**Key Insight:**

The function $h(q)$ represents a standardized effect that is independent of trait measurement units. This allows transfer of information between:

- Continuous trait $\rightarrow$ Binary trait
- Binary trait $\rightarrow$ Continuous trait
- Different continuous traits with different units

**When to Use:**

- Training and target traits differ (e.g., BMD $\rightarrow$ osteoporosis)
- Only p-values available (no BETA values)
- You want to leverage large studies with only summary statistics
- Cross-trait prediction is scientifically justified

### Method Selection Logic

The `method` parameter controls which estimation approach to use:

```{r method-logic, eval=FALSE}
# method = "auto" (default)
# Automatically chooses:
#   - "both" if same trait AND BETA available
#   - "pvalue" otherwise

# method = "beta"
# Direct beta method only (requires BETA)

# method = "pvalue"
# P-value/Z-score method only (requires P, N)

# method = "both"
# Run both methods for comparison (when applicable)
```

**Decision Tree:**

```
Training trait == Target trait?
├─ Yes → BETA available?
│        ├─ Yes → Use "both" (direct beta + p-value)
│        └─ No → Use "pvalue"
└─ No → Use "pvalue" only
```

## Basic Usage Examples

### Example 1: Binary Trait, Same Training/Target, Beta Method

Use direct beta method when you have effect sizes from the same trait:

```{r example1, eval=FALSE}
# Simulate training data (or load from literature)
set.seed(123)
training_MAF <- runif(100, 0.001, 0.4)
# Effect size decreases with MAF (typical pattern)
training_BETA <- 0.3 * sqrt(training_MAF * (1 - training_MAF)) + rnorm(100, 0, 0.05)

# Your study variants
target_MAF <- c(0.01, 0.05, 0.1, 0.2, 0.3)

# Estimate B using direct beta method
B_beta <- get_B(
  training_trait = "binary",
  training_MAF = training_MAF,
  training_BETA = training_BETA,
  target_trait = "binary",
  target_MAF = target_MAF,
  method = "beta"  # Direct beta method only
)

print(data.frame(MAF = target_MAF, B = B_beta))
```

### Example 2: Binary Trait, P-value Method

When you only have p-values (no effect sizes):

```{r example2, eval=FALSE}
# Training data with p-values
set.seed(456)
training_MAF <- runif(100, 0.001, 0.4)
training_P <- runif(100, 1e-8, 0.1)
training_N <- sample(1000:10000, 100, replace = TRUE)

# Target parameters
target_MAF <- seq(0.01, 0.4, by = 0.05)
target_case_prop <- 0.1  # 10% cases in target case-control study

# Estimate B using p-value method
B_pval <- get_B(
  training_trait = "binary",
  training_MAF = training_MAF,
  training_P = training_P,
  training_N = training_N,
  target_trait = "binary",
  target_MAF = target_MAF,
  target_case_prop = target_case_prop,
  method = "pvalue"
)

print(head(data.frame(MAF = target_MAF, B = B_pval)))
```

### Example 3: Continuous Trait

For continuous traits (e.g., height, BMI), you need standard errors:

```{r example3, eval=FALSE}
# Training data from continuous trait GWAS
set.seed(789)
training_MAF <- runif(80, 0.01, 0.5)
training_P <- runif(80, 1e-6, 0.05)
training_N <- rep(10000, 80)

# Target continuous trait with SE
target_MAF <- seq(0.05, 0.45, by = 0.05)
target_SE <- rep(0.1, length(target_MAF))  # Standard error in target study

# Estimate B
B_continuous <- get_B(
  training_trait = "continuous",
  training_MAF = training_MAF,
  training_P = training_P,
  training_N = training_N,
  target_trait = "continuous",
  target_MAF = target_MAF,
  target_SE = target_SE,
  method = "pvalue"
)

print(head(data.frame(MAF = target_MAF, B = B_continuous)))
```

### Example 4: Cross-Trait Estimation

Transfer information from continuous trait (BMD) to binary trait (osteoporosis):

```{r example4, eval=FALSE}
# Training data from BMD GWAS (continuous trait)
set.seed(101)
training_MAF <- runif(200, 0.01, 0.5)
training_P <- runif(200, 1e-10, 0.01)
training_N <- sample(50000:100000, 200, replace = TRUE)

# Target is binary osteoporosis outcome
target_MAF <- seq(0.01, 0.5, length.out = 50)
target_case_prop <- 0.20  # 20% osteoporosis cases in study

# Cross-trait estimation
B_cross <- get_B(
  training_trait = "continuous",  # BMD is continuous
  training_MAF = training_MAF,
  training_P = training_P,
  training_N = training_N,
  target_trait = "binary",  # Osteoporosis is binary
  target_MAF = target_MAF,
  target_case_prop = target_case_prop,
  method = "pvalue",  # Only p-value method works for cross-trait
  verbose = 2
)

# Plot the transfer
plot(target_MAF, B_cross, type = "l", lwd = 2, col = "purple",
     xlab = "MAF", ylab = "Estimated B (for binary outcome)",
     main = "Cross-trait: BMD (continuous) to Osteoporosis (binary)")
```

## Working with Real Data

### Using ALS Example Dataset

The package includes processed ALS (Amyotrophic Lateral Sclerosis) data from literature studies:

```{r als-example, eval=FALSE}
library(GLOWr)

# Load ALS data (lazy-loaded package data)
data(ALS_snvs_B_training)

# Examine the data
str(ALS_snvs_B_training)
head(ALS_snvs_B_training)

# Training data components
training_MAF <- ALS_snvs_B_training$MAF
training_BETA <- ALS_snvs_B_training$BETA
training_P <- ALS_snvs_B_training$P
training_N <- ALS_snvs_B_training$N

# Remove any missing values
valid_idx <- complete.cases(data.frame(training_MAF, training_BETA, training_P, training_N)) &
             training_MAF > 0 & training_MAF < 1 &
             abs(training_BETA) > 1e-10

training_MAF <- training_MAF[valid_idx]
training_BETA <- training_BETA[valid_idx]
training_P <- training_P[valid_idx]
training_N <- training_N[valid_idx]

cat(sprintf("ALS training data: %d variants after QC\n", length(training_MAF)))

# Define target MAFs (e.g., variants in your ALS study)
target_MAF <- seq(0.001, 0.4, length.out = 100)

# ALS case proportion (rare disease)
# This is the proportion of cases in your case-control study
# NOT the population prevalence
target_case_prop <- 0.001  # Example: 200 cases, 200,000 controls

# Estimate B with both methods for comparison
B_als <- get_B(
  training_trait = "binary",
  training_MAF = training_MAF,
  training_BETA = training_BETA,
  training_P = training_P,
  training_N = training_N,
  target_trait = "binary",
  target_MAF = target_MAF,
  target_case_prop = target_case_prop,
  method = "both",  # Run both for comparison
  return_full = TRUE,
  verbose = 2
)

# View method comparison
# The model is nested under $model (a glow_B_model object)
print(B_als$model$method_used)
if (!is.null(B_als$model$comparison$prediction)) {
  cat("\nMethod Comparison (on predictions):\n")
  cat(sprintf("  Correlation: %.3f\n", B_als$model$comparison$prediction$correlation))
  cat(sprintf("  Mean %% difference: %.1f%%\n",
              B_als$model$comparison$prediction$mean_percent_diff))
}

# Plot both methods
# B predictions are stored as B_beta_method / B_pvalue_method
par(mfrow = c(1, 2))
plot(target_MAF, B_als$B_beta_method, type = "l", lwd = 2, col = "blue",
     xlab = "MAF", ylab = "B", main = "Direct Beta Method")
plot(target_MAF, B_als$B_pvalue_method, type = "l", lwd = 2, col = "red",
     xlab = "MAF", ylab = "B", main = "P-value/Z-score Method")
par(mfrow = c(1, 1))

# Use primary estimates
B_final <- B_als$B
```

### Using BMD Example Dataset

Example with continuous trait (Bone Mineral Density):

```{r bmd-example, eval=FALSE}
# Load BMD data (lazy-loaded package data)
data(BMD_snvs_B_training)

# Examine data
str(BMD_snvs_B_training)
head(BMD_snvs_B_training)

# Training data
training_MAF <- BMD_snvs_B_training$MAF
training_P <- BMD_snvs_B_training$P
training_N <- BMD_snvs_B_training$N

# QC
valid_idx <- complete.cases(data.frame(training_MAF, training_P, training_N)) &
             training_MAF > 0 & training_MAF < 1 &
             training_P > 0 & training_P < 1

training_MAF <- training_MAF[valid_idx]
training_P <- training_P[valid_idx]
training_N <- training_N[valid_idx]

cat(sprintf("BMD training data: %d variants after QC\n", length(training_MAF)))

# Target parameters
target_MAF <- seq(0.01, 0.5, length.out = 100)

# For binary osteoporosis outcome
target_case_prop <- 0.20  # 20% cases in study

# Estimate B (cross-trait: continuous to binary)
B_bmd <- get_B(
  training_trait = "continuous",
  training_MAF = training_MAF,
  training_P = training_P,
  training_N = training_N,
  target_trait = "binary",
  target_MAF = target_MAF,
  target_case_prop = target_case_prop,
  method = "pvalue",  # P-value method for cross-trait
  return_full = TRUE,
  verbose = 2
)

# Plot (B_pvalue_method is the same as B when only p-value method was used)
plot(target_MAF, B_bmd$B, type = "l", lwd = 2, col = "darkgreen",
     xlab = "MAF", ylab = "B (binary outcome)",
     main = "BMD to Osteoporosis: Cross-trait B Estimation")
```

### Comparing Both Methods with `method="both"`

When both methods are applicable, compare them to validate estimates:

```{r compare-methods, eval=FALSE}
# Use ALS data (same trait, both BETA and P available)
data(ALS_snvs_B_training)

# Prepare data
training_MAF <- ALS_snvs_B_training$MAF
training_BETA <- ALS_snvs_B_training$BETA
training_P <- ALS_snvs_B_training$P
training_N <- ALS_snvs_B_training$N

# Remove invalid
valid_idx <- complete.cases(data.frame(training_MAF, training_BETA, training_P, training_N)) &
             training_MAF > 0 & training_MAF < 1
training_MAF <- training_MAF[valid_idx]
training_BETA <- training_BETA[valid_idx]
training_P <- training_P[valid_idx]
training_N <- training_N[valid_idx]

# Target
target_MAF <- seq(0.01, 0.4, length.out = 50)

# Run with both methods
result <- get_B(
  training_trait = "binary",
  training_MAF = training_MAF,
  training_BETA = training_BETA,
  training_P = training_P,
  training_N = training_N,
  target_trait = "binary",
  target_MAF = target_MAF,
  target_case_prop = 0.001,
  method = "both",  # Run both methods
  return_full = TRUE,
  verbose = 2
)

# Compare results
# Prediction-based comparison is in result$model$comparison$prediction
pred_comp <- result$model$comparison$prediction
cat("\n=== Method Comparison ===\n")
cat(sprintf("Correlation: %.4f\n", pred_comp$correlation))
cat(sprintf("RMSE: %.4f\n", pred_comp$rmse))
cat(sprintf("Mean percent diff: %.2f%%\n", pred_comp$mean_percent_diff))
cat(sprintf("Max percent diff: %.2f%%\n", pred_comp$max_percent_diff))

# Visualize comparison
par(mfrow = c(2, 2))

# Scatterplot (note: B_beta_method / B_pvalue_method field names)
plot(result$B_beta_method, result$B_pvalue_method,
     xlab = "Direct Beta Method", ylab = "P-value Method",
     main = "Method Comparison", pch = 16, col = rgb(0,0,1,0.5))
abline(0, 1, col = "red", lwd = 2, lty = 2)
legend("topleft", legend = sprintf("r = %.3f", pred_comp$correlation),
       bty = "n")

# Both methods vs MAF
plot(target_MAF, result$B_beta_method, type = "l", lwd = 2, col = "blue",
     xlab = "MAF", ylab = "B", main = "Both Methods",
     ylim = range(c(result$B_beta_method, result$B_pvalue_method)))
lines(target_MAF, result$B_pvalue_method, lwd = 2, col = "red", lty = 2)
legend("topright", legend = c("Beta", "P-value"),
       col = c("blue", "red"), lty = c(1, 2), lwd = 2, bty = "n")

# Difference vs MAF
diff <- result$B_beta_method - result$B_pvalue_method
plot(target_MAF, diff, type = "l", lwd = 2, col = "purple",
     xlab = "MAF", ylab = "Difference (Beta - Pvalue)",
     main = "Method Differences")
abline(h = 0, col = "gray", lty = 2)

# Percent difference
pct_diff <- 100 * abs(diff) / ((result$B_beta_method + result$B_pvalue_method) / 2)
plot(target_MAF, pct_diff, type = "l", lwd = 2, col = "orange",
     xlab = "MAF", ylab = "Percent Difference (%)",
     main = "Percent Differences")
abline(h = 10, col = "red", lty = 2)

par(mfrow = c(1, 1))

# Decide which to use
# If correlation > 0.9 and mean % diff < 10%: Either method is fine
# If correlation > 0.7 and mean % diff < 20%: Reasonable agreement
# Otherwise: Investigate discrepancies
if (pred_comp$correlation > 0.9) {
  cat("\n[OK] High correlation between methods - both are reliable\n")
  cat("Recommendation: Use direct beta method (primary)\n")
} else if (pred_comp$correlation > 0.7) {
  cat("\n[NOTE] Moderate correlation between methods\n")
  cat("Recommendation: Investigate differences, consider averaging\n")
} else {
  cat("\n[WARNING] Low correlation between methods\n")
  cat("Recommendation: Check data quality and parameter specifications\n")
}
```

## Diagnostics and Model Selection

### Using `return_full=TRUE`

Get detailed results including fitted models:

```{r diagnostics-returnfull, eval=FALSE}
# Run with full output
result <- get_B(
  training_trait = "binary",
  training_MAF = training_MAF,
  training_BETA = training_BETA,
  target_trait = "binary",
  target_MAF = target_MAF,
  method = "beta",
  return_full = TRUE  # Get full result object
)

# Access components
names(result)
# [1] "B"              "B_beta_method"  "B_pvalue_method"
# [4] "model"          "target_summary"

# The fitted models are nested under result$model$models
# result$model is a glow_B_model object
names(result$model)
# [1] "method_used"      "models"           "all_models_info"
# [4] "outliers"         "training_summary"  "training_data"
# [7] "comparison"       "selection_criterion"

# Examine fitted model for the beta method
summary(result$model$models$beta_method)

# Check which transformation was selected
formula(result$model$models$beta_method)

# Model fit quality
cat(sprintf("R-squared: %.4f\n",
    summary(result$model$models$beta_method)$r.squared))
cat(sprintf("Adjusted R-squared: %.4f\n",
    summary(result$model$models$beta_method)$adj.r.squared))
```

### Plotting Results with `plot()`

Visualize model fit and diagnostics:

```{r diagnostics-plot, eval=FALSE}
# Get full result
result <- get_B(
  training_trait = "binary",
  training_MAF = training_MAF,
  training_BETA = training_BETA,
  training_P = training_P,
  training_N = training_N,
  target_trait = "binary",
  target_MAF = target_MAF,
  target_case_prop = 0.001,
  method = "both",
  return_full = TRUE
)

# Dashboard view (2x2 grid)
plot(result)  # Equivalent to plot(result, type = "all")

# Individual plot types
plot(result, type = "fit")         # Model fit
plot(result, type = "residuals")   # Residual diagnostics
plot(result, type = "comparison")  # Method comparison (if both methods)
```

The dashboard includes:

- **Top-left**: Training data vs fitted curve (shows model quality)
- **Top-right**: Q-Q plot (checks residual normality)
- **Bottom-left**: Residuals vs fitted (checks for patterns)
- **Bottom-right**: Method comparison or Cook's distance

### Running Diagnostics with `get_B_diagnostics()`

Comprehensive diagnostic report:

```{r diagnostics-detailed, eval=FALSE}
# Get full result first
result <- get_B(
  training_trait = "binary",
  training_MAF = training_MAF,
  training_BETA = training_BETA,
  training_P = training_P,
  training_N = training_N,
  target_trait = "binary",
  target_MAF = target_MAF,
  target_case_prop = 0.001,
  method = "both",
  return_full = TRUE
)

# Run diagnostics
diag <- get_B_diagnostics(
  result,
  show_plots = TRUE,      # Display diagnostic plots
  outlier_threshold = 4,  # Cook's distance threshold
  verbose = 1             # Print summary
)

# Examine components
names(diag)
# [1] "model_fit"         "residual_stats"    "outliers"
# [4] "method_comparison" "qc_flags"          "summary_text"

# Model fit statistics
diag$model_fit$beta$R2       # R-squared
diag$model_fit$beta$adj_R2   # Adjusted R-squared
diag$model_fit$beta$formula  # Selected model

# Residual statistics
diag$residual_stats$mean         # Should be ~0
diag$residual_stats$normality_p  # Shapiro-Wilk p-value

# Outliers
diag$outliers$n_outliers     # Number of outliers
diag$outliers$proportion     # Proportion of data
diag$outliers$indices        # Which observations

# Quality control flags
if (length(diag$qc_flags) > 0) {
  cat("\nQuality Control Issues:\n")
  for (flag in diag$qc_flags) {
    cat("  [!]", flag, "\n")
  }
} else {
  cat("\n[OK] No quality control issues detected\n")
}
```

### Interpreting Diagnostic Output

**Model Fit (R²):**

- **R² > 0.5**: Good fit - model captures MAF-effect relationship well
- **R² 0.3-0.5**: Moderate fit - acceptable but more data may help
- **R² < 0.3**: Poor fit - warning issued, consider data quality

**Residuals:**

- **Mean ≈ 0**: Model is unbiased
- **Normal distribution**: Model assumptions met
- **Non-normal** (p < 0.05): May indicate outliers or model misspecification

**Outliers (Cook's Distance):**

- Cook's D > 4/n flags potentially influential points
- **<10% outliers**: Typical, usually not problematic
- **>10% outliers**: Warning issued, investigate data quality

**Method Comparison:**

- **Correlation > 0.9**: Excellent agreement
- **Correlation 0.7-0.9**: Good agreement
- **Correlation < 0.7**: Warning issued, investigate discrepancies

## Advanced Topics

### Understanding the 8 Candidate Models

The function tests 8 different transformations of X (MAF) and Y (effect size measure):

**X transformations:**

1. $X = q$ (raw MAF)
2. $f(X) = q(1-q)$ (heterozygosity/frequency function)
3. $\log(X) = \log(q)$ (log MAF)
4. $\log[f(X)] = \log[q(1-q)]$ (log frequency function)

**Y transformations:**

1. $Y$ (raw squared effect)
2. $\log(Y)$ (log of squared effect)

**Why multiple models?**

Different genetic architectures produce different MAF-effect relationships:

- **Rare variants with large effects**: $\log(Y) \sim \log(X)$ often fits well
- **Common variants**: $Y \sim X$ or $Y \sim f(X)$ may be better
- **Complex patterns**: Log transformations capture exponential relationships

The function automatically selects the best-fitting model by R², but you can compare all models:

```{r advanced-models, eval=FALSE}
# Get full result with fitted model
result <- get_B(
  training_trait = "binary",
  training_MAF = training_MAF,
  training_BETA = training_BETA,
  target_trait = "binary",
  target_MAF = target_MAF,
  return_full = TRUE
)

# See which model was selected
cat("Selected model formula:\n")
print(formula(result$model$models$beta_method))

# Check R-squared
cat(sprintf("\nR-squared: %.4f\n",
    summary(result$model$models$beta_method)$r.squared))

# Note: compare_B_models() has limited functionality in current implementation
# It can only show the selected model, not re-fit all candidates
# This is documented in the function's limitations
```

### When Estimates Differ Between Methods

If direct beta and p-value methods give different results:

**Common Causes:**

1. **Different training data used**: Beta method uses different variants than p-value method
2. **Target parameters**: Check `target_case_prop` or `target_SE` are correct
3. **Sample size heterogeneity**: P-value method sensitive to varying N across studies
4. **Winner's curse**: Published effect sizes may be inflated

**What to do:**

```{r estimates-differ, eval=FALSE}
# If methods disagree, investigate:

# 1. Check correlation (prediction comparison is under $model$comparison$prediction)
pred_comp <- result$model$comparison$prediction
if (pred_comp$correlation < 0.7) {
  cat("Low correlation - investigate further\n")

  # 2. Plot difference vs MAF
  plot(target_MAF, result$B_beta_method - result$B_pvalue_method,
       xlab = "MAF", ylab = "Difference", main = "Method Differences by MAF")
  abline(h = 0, col = "red", lty = 2)
  # Are differences systematic or random?

  # 3. Check outliers
  diag <- get_B_diagnostics(result, show_plots = FALSE)
  if (diag$outliers$n_outliers > 0.1 * length(training_MAF)) {
    cat("Many outliers detected - may affect fit\n")
  }

  # 4. Check model fit quality (models are in result$model$models)
  cat(sprintf("Beta method R²: %.3f\n",
      summary(result$model$models$beta_method)$r.squared))
  cat(sprintf("Pvalue method R²: %.3f\n",
      summary(result$model$models$pvalue_method)$r.squared))
  # If one R² is much lower, that method may be unreliable
}

# Decision rule:
# - High correlation + similar R²: Use either (prefer beta method)
# - Low correlation + similar R²: Average the two
# - Low correlation + different R²: Use method with higher R²
# - Very low correlation: Investigate data quality issues
```

### Quality Control Considerations

**Training Data QC:**

Before running `get_B()`, ensure:

```{r qc-training, eval=FALSE}
# Check MAF range
summary(training_MAF)
stopifnot(all(training_MAF > 0 & training_MAF <= 0.5))

# Check for missing values
stopifnot(all(!is.na(training_MAF)))
stopifnot(all(!is.na(training_BETA)))

# Check effect sizes are reasonable
hist(abs(training_BETA), breaks = 30, main = "Distribution of |BETA|")
# Flag extremely large effects (potential errors)
if (any(abs(training_BETA) > 2)) {
  warning("Some effect sizes |BETA| > 2 - check for data errors")
}

# Check sample sizes
if (!is.null(training_N)) {
  summary(training_N)
  # Recommend N > 500 for reliable estimates
  if (any(training_N < 500)) {
    warning(sprintf("%d variants from studies with N < 500", sum(training_N < 500)))
  }
}
```

**Target Data QC:**

```{r qc-target, eval=FALSE}
# Ensure MAF in valid range
stopifnot(all(target_MAF > 0 & target_MAF <= 0.5))

# Check target parameters are sensible
if (!is.null(target_case_prop)) {
  stopifnot(all(target_case_prop > 0 & target_case_prop < 1))
  # For case-control: should be case proportion in STUDY, not population
  if (any(target_case_prop < 0.001)) {
    warning("Very low case proportions - verify this is case prop in study, not population prevalence")
  }
}

if (!is.null(target_SE)) {
  stopifnot(all(target_SE > 0))
  # SE should be on same scale as training data
}
```

### Handling Edge Cases

**Very rare variants (MAF < 0.001):**

```{r edge-rare, eval=FALSE}
# Extrapolation to very rare MAF may be unreliable
target_MAF_rare <- c(0.0001, 0.0005, 0.001)

# Check if training data includes similarly rare variants
min_training_MAF <- min(training_MAF)
cat(sprintf("Minimum training MAF: %.5f\n", min_training_MAF))

if (min(target_MAF_rare) < min_training_MAF) {
  warning("Extrapolating to MAF rarer than training data - estimates may be unreliable")
  # Consider: Use only target MAF within training range
  target_MAF_safe <- target_MAF_rare[target_MAF_rare >= min_training_MAF]
}
```

**Common variants (MAF > 0.4):**

```{r edge-common, eval=FALSE}
# Similar issue at upper end
max_training_MAF <- max(training_MAF)
cat(sprintf("Maximum training MAF: %.3f\n", max_training_MAF))

# If target includes MAF > max training, consider limiting
target_MAF_safe <- target_MAF[target_MAF <= max_training_MAF]
```

**Small training sample:**

```{r edge-small, eval=FALSE}
# With < 20 variants, model may overfit
if (length(training_MAF) < 20) {
  warning("Small training sample (N < 20) - consider simpler model")
  # Simpler models: prefer Y ~ X or Y ~ f(X)
  # Current implementation always tests all 8 models
}

# Recommendation: ≥ 50 variants for reliable fit
if (length(training_MAF) < 50) {
  cat("Note: Training N =", length(training_MAF), "- more data recommended\n")
}
```

## Practical Recommendations

### Best Practices for Data Preparation

**1. Harmonize MAF:**

```{r best-harmonize, eval=FALSE}
# Ensure MAF ≤ 0.5 (not > 0.5)
# If you have risk allele frequency (RAF), convert:
harmonize_MAF <- function(freq) {
  ifelse(freq > 0.5, 1 - freq, freq)
}

training_MAF <- harmonize_MAF(training_RAF)
```

**2. Remove duplicates:**

```{r best-duplicates, eval=FALSE}
# If multiple studies report same variant, keep largest study
if (anyDuplicated(training_rsID)) {
  # Order by sample size
  training_data <- training_data[order(training_data$N, decreasing = TRUE), ]
  # Remove duplicates
  training_data <- training_data[!duplicated(training_data$rsID), ]
}
```

**3. Apply sample size filter:**

```{r best-samplesize, eval=FALSE}
# Exclude small studies (N < 500)
min_N <- 500
keep_idx <- training_N >= min_N
training_MAF <- training_MAF[keep_idx]
training_BETA <- training_BETA[keep_idx]
training_P <- training_P[keep_idx]
training_N <- training_N[keep_idx]

cat(sprintf("Retained %d/%d variants with N >= %d\n",
            sum(keep_idx), length(keep_idx), min_N))
```

### Choosing Between Methods

**Use Direct Beta Method when:**

- Training and target traits are identical (e.g., both ALS)
- Effect sizes (BETA) available
- You want the most direct relationship
- Training data is well-powered (N > 1000)

**Use P-value Method when:**

- Only p-values available (no BETA)
- Cross-trait prediction (e.g., BMD → osteoporosis)
- Training data from multiple studies with different scales
- You want to leverage very large studies

**Use Both Methods when:**

- Same trait type AND both BETA and p-values available
- You want to validate estimates
- Preparing results for publication
- Assessing robustness

### Interpreting Results

**Effect Size Magnitude:**

```{r interpret-magnitude, eval=FALSE}
# For binary traits (log odds ratios):
# |B| < 0.1: Small effect
# |B| 0.1-0.3: Moderate effect
# |B| > 0.3: Large effect

# For continuous traits (standardized):
# |B| < 0.2: Small effect
# |B| 0.2-0.5: Moderate effect
# |B| > 0.5: Large effect

# Check your results
summary(B_estimates)
hist(B_estimates, breaks = 30, main = "Distribution of Estimated B")

# Flag extreme values
extreme <- B_estimates > quantile(B_estimates, 0.99)
if (any(extreme)) {
  cat(sprintf("%d extreme B values detected (top 1%%)\n", sum(extreme)))
  cat("MAFs with extreme B:", target_MAF[extreme], "\n")
}
```

**MAF-Effect Relationship:**

```{r interpret-relationship, eval=FALSE}
# Typical patterns:
# 1. Decreasing: B decreases as MAF increases (purifying selection)
# 2. U-shaped: Lowest B at intermediate MAF
# 3. Constant: B doesn't vary with MAF (neutral)

# Plot to visualize
plot(target_MAF, B_estimates, type = "l", lwd = 2,
     xlab = "MAF", ylab = "B",
     main = "Effect Size vs MAF Pattern")

# Check monotonicity
diffs <- diff(B_estimates)
if (all(diffs < 0)) {
  cat("Pattern: Monotonically decreasing (typical for deleterious variants)\n")
} else if (all(diffs > 0)) {
  cat("Pattern: Monotonically increasing (unusual - check data)\n")
} else {
  cat("Pattern: Non-monotonic\n")
}
```

### Common Pitfalls to Avoid

**1. Confusing case proportion with prevalence:**

```{r pitfall-prevalence, eval=FALSE}
# WRONG: Using population prevalence
target_case_prop <- 0.00015  # ALS population prevalence
# This is NOT what you want!

# CORRECT: Using study case proportion
# If you have 200 cases and 200 controls:
n_cases <- 200
n_controls <- 200
target_case_prop <- n_cases / (n_cases + n_controls)  # = 0.5
# This is the correct value to use
```

**2. Extrapolating beyond training range:**

```{r pitfall-extrapolate, eval=FALSE}
# Check training range
range(training_MAF)
# [1] 0.01 0.45

# AVOID: Target MAF outside this range
target_MAF_bad <- seq(0.0001, 0.5, length.out = 100)  # Goes to 0.0001

# BETTER: Stay within training range
target_MAF_good <- seq(0.01, 0.45, length.out = 100)
```

**3. Using wrong standard error:**

```{r pitfall-se, eval=FALSE}
# For continuous traits with p-value method:
# SE should match the trait scale

# WRONG: Using SE from different trait
target_SE_wrong <- 0.001  # SE from height study
# Applied to BMI estimation - scales don't match!

# CORRECT: Use SE from target trait or similar trait
target_SE_correct <- 0.05  # SE appropriate for BMI scale
```

**4. Ignoring diagnostics:**

```{r pitfall-diagnostics, eval=FALSE}
# DON'T just take B estimates blindly
B <- get_B(...)  # Get estimates

# DO check quality
result <- get_B(..., return_full = TRUE)
diag <- get_B_diagnostics(result)

if (length(diag$qc_flags) > 0) {
  cat("WARNING: Quality issues detected!\n")
  # Investigate before using estimates
}
```

## Summary and Recommendations

### Quick Reference Guide

**For beginners:**

```{r summary-beginner, eval=FALSE}
# Simplest approach - automatic method selection
B <- get_B(
  training_trait = "binary",
  training_MAF = training_MAF,
  training_BETA = training_BETA,  # Or use training_P if no BETA
  target_trait = "binary",
  target_MAF = target_MAF,
  target_case_prop = 0.1  # Adjust for your study
)
```

**For production analysis:**

```{r summary-production, eval=FALSE}
# Full workflow with QC and diagnostics
result <- get_B(
  training_trait = "binary",
  training_MAF = training_MAF,
  training_BETA = training_BETA,
  training_P = training_P,
  training_N = training_N,
  target_trait = "binary",
  target_MAF = target_MAF,
  target_case_prop = target_case_prop,
  method = "both",       # Compare methods
  return_full = TRUE,    # Get full diagnostics
  verbose = 2            # Informative output
)

# Check quality
diag <- get_B_diagnostics(result, show_plots = TRUE)
if (length(diag$qc_flags) == 0) {
  cat("[OK] Ready for downstream analysis\n")
  B_final <- result$B
} else {
  cat("[WARNING] Review quality issues before proceeding\n")
}
```

### When to Use Which Method

| Scenario | Method | Notes |
|----------|--------|-------|
| Same trait, BETA available | `method="beta"` or `"both"` | Direct beta preferred |
| Same trait, no BETA | `method="pvalue"` | Only option |
| Different traits | `method="pvalue"` | Only option |
| Validation needed | `method="both"` | Compare for robustness |
| Publication | `method="both"` | Show consistency |

### Minimum Requirements

**Training data:**

- **N variants**: ≥ 50 recommended (20 minimum)
- **Sample sizes**: ≥ 500 per study
- **MAF range**: Should overlap with target MAF

**Target data:**

- **MAF range**: Within or close to training range
- **Parameters**: Correctly specified (case_prop or SE)

### Integration with GLOW Pipeline

Use estimated B values in optimal weight calculation:

```{r summary-integration, eval=FALSE}
# 1. Estimate B
B <- get_B(
  training_trait = "binary",
  training_MAF = training_MAF,
  training_BETA = training_BETA,
  target_trait = "binary",
  target_MAF = target_MAF,
  target_case_prop = 0.1
)

# 2. Get variant-importance scores (PI)
# (Assume you have these from annotation tools)
PI <- c(0.1, 0.3, 0.05, 0.8, 0.2, 0.4, ...)  # One per variant

# 3. Get correlation matrix M
# (From LD or kinship matrix)
M <- cor(genotype_matrix)  # Or from LD reference

# 4. Calculate optimal weights
g_identity <- function(x) x
weights <- Optimal_Weights_M(g_identity, B, PI, M)

# 5. Use weights in association test
# (See other vignettes for GLOW test functions)
```

## See Also

**Related Functions:**

- `get_B()` - All-in-one effect size estimation (train + predict)
- `train_B_model()` - Train model separately (returns `glow_B_model`)
- `predict_B()` - Predict B from a trained model
- `save_B_model()` / `load_B_model()` - Persist and reload trained models
- `plot.glow_B_estimate()` / `plot.glow_B_model()` - Visualization methods
- `get_B_diagnostics()` - Comprehensive diagnostic report
- `compare_B_models()` - Model comparison tools
- `select_best_model()` - Internal model selection

**Related Vignettes:**

- "Diagnostic Tools for B Estimation" - Detailed diagnostics
- "Optimal Weight Calculation" - Using B in GLOW framework
- "Complete GLOW Pipeline" - End-to-end workflow

**References:**

Zhang, H., Liu, M., Landers, J. E., & Wu, Z. Integrated Weighted Association Test with Application to Genetic Association Studies. *Annals of Applied Statistics* (in revision).

## Session Information

For reproducibility:

```{r sessioninfo, eval=TRUE}
sessionInfo()
```
