---
title: "Quick start: annotating with FAVOR"
author: "GLOWr Package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick start: annotating with FAVOR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This vignette is the short route through `annotate_favor()`: when to use it, what to have
ready, which settings to choose, what comes out, and how to check it. The companion vignette
"Annotating genotype data with FAVOR" (`vignette("favor-annotation", package = "GLOWr")`) gives
the evidence, the cost measurements and the matching procedure in full. 

## When to use annotate_favor()

Use `annotate_favor()` when you have variants on GRCh38 and need FAVOR's functional
annotation attached to them: a cohort GDS that a GLOW scan (the variant-set association
method of this package, run by the GLOWpipeline package) or STAARpipeline will read, or a
table of variants for training GLOWr's variant-importance score (PI), which weights the
variants in a scan. The function reads the FAVOR database from
your own disk, finds the database row for each variant, and returns the annotation together
with a record of how each variant was matched. With `output_agds` it also writes the annotated
GDS (an aGDS) in the layout that GLOW's scans and STAARpipeline read.

FAVOR keys every variant by `CHR-POS-REF-ALT`, where `REF` is the base the reference genome
carries. `match_method = "exact"` looks the key up exactly as given, so it annotates only
the variants whose alleles already follow that convention. Sequencing data in VCF form do. Genotyping-chip data usually do not, because a PLINK file carries each probe's
two alleles in the probe's own order and on the probe's strand. Beyond the exact lookup,
`annotate_favor()` adds four things:

- **Flexible allele matching, with a check.** For a chip SNV it also tries the key with the
  alleles swapped, with both complemented, and with both, and it records which form found
  the row (`match_tier`). It then compares the input's rsID with the matched row's
  (`rsid_check`) and by default withholds a rewritten match that the rsID contradicts. Your
  genotype file is never changed. On an ALS case-control data set genotyped on an Illumina array, about 48,000
  samples and 413,528 variants, the allele labeled `REF` was the reference base for 36% of
  the variants, so exact matching would have annotated that share only. Flexible matching
  found a row for all of them, and the chip's rsID agreed with the FAVOR row's for 99.89% of
  the 265,462 matches found by a rewritten key.

- **Variants without alleles.** `annotate_favor()` annotates a key of the form
  `CHR-POS-NA-NA`, common in literature-curated variant lists, from the SNV rows at the
  position, and labels it `position_only`, so you know which values are aggregates.

- **Both database releases, read locally.** The FAVOR v1 CSV chunks and the FAVOR 2.0 Parquet
  files are both read, and the provenance of the annotation (database files, settings, outcome
  counts, date) is written into the aGDS.

- **Throughput for a whole cohort.** With FAVOR 2.0, the 22 autosomes of that chip data set
  took 2,333 seconds in total, and chromosome 22 of the 1000 Genomes high-coverage sequencing
  panel (3,202 samples, 1,066,557 variants) took 205 seconds. The companion vignette's cost
  table gives time and peak memory for both database routes.

## What you need

`annotate_favor()` needs these ready before the call: variants on GRCh38 in one of three
forms, the FAVOR database on disk, two R packages, rsIDs when the input is chip data, and disk
space for the aGDS.

| Need | Details |
| --- | --- |
| Variants on GRCh38 | FAVOR's positions are GRCh38. Convert the positions first if your data are on another build. |
| The variants, in one of three forms | A GDS file path (one chromosome per call, the usual form for a cohort). A data frame with a `VarInfo` column of `CHR-POS-REF-ALT` keys and, optionally, an `rsID` column. The case or control object from `prepare_PI_case_data()` or `prepare_PI_control_data()`. |
| The FAVOR database on disk | Obtained from the FAVOR team, since GLOWr does not redistribute it. Either the v1 CSV chunk directory of `chr<N>_<K>.csv` files (its split table `FAVORdatabase_chrsplit.csv` is bundled with GLOWr), or the FAVOR 2.0 directory of `chromosome_<N>.parquet` files. |
| R packages | GLOWr. `SeqArray` for GDS input and aGDS output. `arrow` for the Parquet files. The optional command-line tool `xsv` speeds up the CSV route. Without it the chunks are read with `data.table::fread()`. |
| rsIDs in the input, for chip data | A GDS made with `SeqArray::seqBED2GDS()` carries column 2 of the `.bim` as `annotation/id`. That is the evidence the rsID check reads, so keep it. |
| Disk space | The aGDS is a copy of the input GDS plus the annotation nodes. |

## Which settings to choose

The call for one chromosome of chip data is one line per argument. Replace the three paths
with your own. The example does not create them.

```{r call}
res <- annotate_favor(
  variants      = "genotypes/chr22.gds",          # your GDS, one chromosome
  favor_db_path = "/path/to/FAVOR2.0",            # the FAVOR 2.0 directory, or the v1 chunk directory
  match_method  = "flexible",                     # "exact" for sequencing data in the reference convention
  favor_release = "FAVOR 2.0, downloaded 2026",   # a free-text label recorded in the aGDS provenance
  output_agds   = "genotypes_favor/chr22.gds",    # the annotated GDS for the scans
  output_csv    = "genotypes_favor/chr22.csv")    # the returned table on disk, for the record
```

The arguments that decide what happens:

| Argument | Choose | Why |
| --- | --- | --- |
| `match_method` | `"flexible"` for chip data. `"exact"` for sequencing data whose alleles follow the reference convention. | Exact matching tries the key as given only. Flexible matching adds the three rewritten keys for SNVs. Indels match by the exact key under both. |
| `rsid_policy` | Leave the default `"require"` for an annotation you will analyze. `"record"` to see what the lookup alone finds. | Under `"require"` a rewritten match whose rsID differs from the row's, or whose row carries no rsID, is withheld as `rsid_conflict`. Exact matches, and inputs without rsIDs, are never withheld. |
| `features` | Leave the default unless you need fewer fields. | The default is the 30 fields of the FAVOR Essential Database: 20 numeric scores and 10 categorical fields, among them the GENCODE and MetaSVM fields the built-in coding masks read. FAVOR 2.0 lacks four older score versions, so the default yields 26 fields there. |
| `output_agds`, `update_gds`, `output_csv` | `output_agds` for a new annotated file. `update_gds = TRUE` to write the annotation into the input GDS instead. `output_csv` for the table. | See the next section. |
| `variant_filter` | `list(chr = "22")`, or `list(chr = "22", start = 1e6, end = 5e6)`, to restrict a call. | One call per chromosome is the natural unit. |
| `favor_db_format` | Leave `"auto"`. | The layout is detected from the directory. A directory holding both layouts, or neither, is an error, and you then name the format. |
| `na_allele_method` | `"average"` (the default) or `"first"`. | Only for `CHR-POS-NA-NA` keys. Resolve the alleles first when you can. |
| `na_handling` | Leave `"keep"`. | `"zero"` replaces a missing numeric value with 0. `"drop"` removes rows with any missing annotation from the table and the CSV. Neither changes the matching outcome. |

## What you get

One call to `annotate_favor()` can produce up to three things: the returned table for
inspection, a CSV copy for the record, and an annotated GDS for the scans of GLOW and
STAARpipeline. The table says which is for auditing and which for analysis.

| Output | What it is | Use it for |
| --- | --- | --- |
| `res`, the returned data frame | The input columns (for a GDS also `variant_id` and `rsID`), one column per annotation feature, and the three records `match_tier`, `favor_key` and `rsid_check`. | Inspecting and auditing the matching. The input of the variant-importance training functions. |
| `output_csv` | The same table written to disk. | The record you keep. With the returned table, the only place `favor_key` is stored. |
| `output_agds`, or the input GDS under `update_gds = TRUE` | A copy of the input GDS plus one node per feature under `annotation/info/FunctionalAnnotation`, the outcome as `annotation/info/favor_match_tier`, the check as `annotation/info/favor_rsid_check`, and the provenance as attributes of the `FunctionalAnnotation` folder. | The downstream analyses: GLOW's gene, window and coding scans, and STAARpipeline. |

`favor_key` is the `CHR-POS-REF-ALT` of the FAVOR row that supplied the annotation, or of the
row withheld under `rsid_conflict`. The aGDS does not store it, so keep the returned table or
request `output_csv` when you need the keys.

## How to know it worked

Two checks confirm that the call worked: the outcome counts in the returned table, and the
same counts read back from the aGDS. Read the outcome table first:

```{r check}
table(res$match_tier)
table(res$rsid_check)
```

- On chip data under flexible matching, expect the four key forms (`exact`, `swapped`,
  `flipped`, `flipped_swapped`), a small number of `rsid_conflict`, and the reasons a key was
  not found. On the chip data set above, the default policy withheld 289 rewritten matches,
  about one in a thousand.

- On sequencing data under exact matching, expect `exact` and the reasons a key was not
  found: `uncovered` (FAVOR has no row at the position), `unmatched_alt` (the position is
  covered and the reference base agrees, but FAVOR has no row for that allele pair) and
  `unmatched_ref` (the input's `REF` is not the reference base).

- `rsid_check` is mostly `same` on chip data. It is `chip_none` throughout when the input
  carries no rsIDs, the usual case for sequencing data, and nothing is then withheld.

- A large `unmatched_ref` share on chip data under exact matching means the alleles do not
  follow the reference convention. Rerun with `match_method = "flexible"`.

Then confirm that the aGDS carries the annotation. The counts must equal those of `res`:

```{r agds}
g <- SeqArray::seqOpen("genotypes_favor/chr22.gds")
table(SeqArray::seqGetData(g, "annotation/info/favor_match_tier"))
table(SeqArray::seqGetData(g, "annotation/info/favor_rsid_check"))
fa <- gdsfmt::index.gdsn(g, "annotation/info/FunctionalAnnotation")
gdsfmt::ls.gdsn(fa)                 # one node per annotation feature
str(gdsfmt::get.attr.gdsn(fa))      # the provenance: database files, settings, counts, date
SeqArray::seqClose(g)
```

## Where to read more

Three places carry more than this quick start does:

- `vignette("favor-annotation", package = "GLOWr")`: the evidence, the cost table, the two
  database layouts, where each record lives, and the optional alignment of chip alleles
  before conversion. Its appendices give every value of `match_tier` and `rsid_check` (A),
  the keys without alleles (B), and a worked example on a made-up database that you can
  run (C).

- `?annotate_favor`: every argument and its default.

- The GLOWanalyses workflow templates run `annotate_favor()` per chromosome as a job array
  (`00-data-prep/annotate-favor.R` with `slurm/run-annotate-favor.sh`) and write a provenance
  snapshot beside the output.

## References

- Zhou, H., Arapoglou, T., Li, X., et al. (2023). FAVOR: functional annotation of variants
  online resource and annotator for variation across the human genome. *Nucleic Acids
  Research*, 51(D1), D1300-D1311. doi:10.1093/nar/gkac966.
- Zhou, H., Verma, V., Li, X., et al. (2026). FAVOR 2.0: A reengineered functional annotation of
  variants online resource for interpreting genomic variation. *Nucleic Acids Research*, 54(D1),
  D1405-D1414. doi:10.1093/nar/gkaf1217.
