---
title: "Annotating genotype data with FAVOR"
author: "GLOWr Package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Annotating genotype data with FAVOR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
has_arrow <- requireNamespace("arrow", quietly = TRUE)
```

## Annotate your genotype data

`annotate_favor()` attaches functional annotation from the FAVOR database to the variants of a
genotype data set and writes the annotated GDS (an aGDS) that GLOW and STAARpipeline scan. FAVOR
keys every variant by `CHR-POS-REF-ALT` on GRCh38, where `REF` is the base the reference genome
carries. Sequencing data in VCF form follow that convention, so their keys name FAVOR's rows
directly. Genotyping-chip data, the usual input of a GWAS, often do not. A PLINK file carries the
two alleles of each probe in the probe's own order and on the strand the probe was designed for,
so the allele labelled `REF` may be the reference base, the other allele, or the complement of
either.

The annotator handles this with flexible matching. For each SNV it looks the key up as given,
then with the two alleles swapped, then with both complemented, then both. It records which form
found the row. It then compares the input's rsID with the rsID of the FAVOR row it found, and
under the default policy it keeps a rewritten match only when the two agree. A rewritten match
is also kept when the input carries no rsID, since nothing then speaks against it. The genotype
data themselves are never modified. Only the lookup key is rewritten, and the aGDS is a copy of
the input plus the annotation.

The evidence that this works comes from an ALS case-control data set genotyped on an Illumina
array, about 48,000 samples and 413,528 variants on chromosomes 1 to 22 after quality control,
converted from PLINK to GDS. There the allele labelled `REF` is the reference base for 36% of the
variants, the allele labelled `ALT` is for 14%, and neither is for 50%. Exact matching would
annotate the first group only. Flexible matching found a row for all of them, and the chip's
rsID agreed with the FAVOR row's for 99.89% of the 265,462 matches found by a rewritten key. The
agreement ranged from 99.83% to 99.92% across the four key forms, the exact matches included.
The default policy withheld the 289 rewritten matches the rsID did not confirm, about one in a
thousand.

The call for a chromosome of chip data is one line per argument. Replace the three paths with
your own. The example does not create them.

```{r gds, eval = FALSE}
res <- annotate_favor(
  variants      = "genotypes/chr22.gds",     # your GDS, one chromosome
  favor_db_path = "/path/to/FAVOR2.0",       # or the FAVOR v1 chunk directory
  match_method  = "flexible",                # rsid_policy = "require" is the default
  output_agds   = "genotypes_favor/chr22.gds")
table(res$match_tier)
table(res$rsid_check)
```

Use `match_method = "flexible"` for chip data. Use `match_method = "exact"` for sequencing data
whose alleles already follow the reference convention. A variant is then annotated only when its
complete key exists in FAVOR. Appendix A gives the procedure in full, with every value that
`match_tier` and `rsid_check` can take.

## Computing costs

The numbers below are the elapsed time of the annotation call, including the CSV and aGDS
output, and the peak resident memory of the R process. They were measured on two data sets with
the default feature set, one call per chromosome:

- the ALS chip data set described above (5,970 of its variants are on chromosome 22).
- the 1000 Genomes Project high-coverage sequencing panel, chromosome 22 only: 3,202 samples
  and 1,066,557 variants, VCF converted to GDS.

For chromosomes 1 to 22 the time is the sum over the calls and the memory is the largest peak
among them. The CSV route's `xsv` child process is not counted in the memory.

| Data | Database | Time | Peak memory |
| --- | --- | --- | --- |
| ALS chip, chr22 (5,970 variants) | FAVOR v1 CSV | 378 s | 0.47 GiB |
| ALS chip, chr22 | FAVOR 2.0 Parquet | 37 s | 2.87 GiB |
| ALS chip, chromosomes 1 to 22 (413,528 variants) | FAVOR v1 CSV | 40,006 s | 0.54 GiB |
| ALS chip, chromosomes 1 to 22 | FAVOR 2.0 Parquet | 2,333 s | 3.61 GiB |
| 1000 Genomes sequencing, chr22 (1,066,557 variants) | FAVOR v1 CSV | 466 s | 4.78 GiB |
| 1000 Genomes sequencing, chr22 | FAVOR 2.0 Parquet | 205 s | 8.44 GiB |

The Parquet route was about 17 times faster over the 22 chip chromosomes, about 10 times faster
on chromosome 22 of the chip data, and about 2 times faster on chromosome 22 of the sequencing
data. The two routes read the database differently. The CSV route streams every chunk that
overlaps the input's position range through `xsv`, which for a chromosome of chip data is every
chunk, so its time is set by the database and barely moves with the number of variants (378 to
466 seconds). The Parquet route reads the position column of every block of rows, and reads the
annotation fields only for the blocks that hold input positions, so its time and memory grow
with the rows it keeps (37 to 205 seconds, 2.87 to 8.44 GiB). The measurements do not separate
reading, matching and output, so they show the pattern and not its cause. The sample count
enters only through the aGDS output, which copies the input GDS before adding the annotation
nodes. The two databases also differ in content, so the rows are comparisons of two routes, not
of two file formats.

## Prepare the FAVOR database

FAVOR is described in Zhou et al. (2023) and FAVOR 2.0 in Zhou et al. (2026), both listed
under References. You obtain the database files from the FAVOR team. GLOWr does not
redistribute them. The annotator reads two layouts:

- **FAVOR v1 CSV**: the per-chromosome chunk files `chr<N>_<K>.csv` with the split table
  `FAVORdatabase_chrsplit.csv`, a copy of which is bundled with GLOWr. The route can use the
  optional command-line tool `xsv` to join the chunks to the input's positions. The tool's
  author has archived its repository, but existing installations work. Without it, GLOWr reads
  the chunks with `data.table::fread()`. Set `use_xsv = FALSE` to choose that route explicitly.
- **FAVOR 2.0 Parquet**: one file `chromosome_<N>.parquet` per chromosome, sorted by position,
  with the annotation in nested fields. It needs the R package `arrow`.

`favor_db_format = "auto"`, the default, detects the layout from the directory. A directory
holding both layouts, or neither, is an error, and you then name the format yourself. Both
readers hand their rows to the same matching procedure. The two releases differ in content,
however, so a variant can match in one and not the other, and its annotation values can differ.

## Inspect the results

The returned table carries the input columns, one column per annotation feature, and three
columns that say what happened to each variant. `match_tier` is the outcome. `favor_key` is the
`CHR-POS-REF-ALT` of the FAVOR row that supplied the annotation, or of the row withheld.
`rsid_check` is the rsID evidence. Read `table(res$match_tier)` first. On chip data you expect
the four key forms and a small number of `rsid_conflict`. On sequencing data you expect `exact`
and the reasons a key was not found.

The aGDS stores the annotation and two of the three records:

| What | Where in the aGDS |
| --- | --- |
| one node per feature | `annotation/info/FunctionalAnnotation/<feature>` |
| the outcome per variant | `annotation/info/favor_match_tier` |
| the rsID check per variant | `annotation/info/favor_rsid_check` |

`favor_key` is not stored in the aGDS. Keep the returned table, or request `output_csv`, when
you need the keys. The `FunctionalAnnotation` folder also carries the provenance as attributes,
read with `gdsfmt::get.attr.gdsn()`. They record:

- the database format, path and files, with their sizes.
- the SHA-256 entries of a `SHA256SUMS` file, when one sits in the database directory or its
  parent.
- the release label you passed as `favor_release`, recorded as `"unknown"` when you did not.
- the GLOWr version, the matching settings, the rsID policy and its source.
- the counts per outcome and per check, and the date.

One call per chromosome is the natural unit, and `variant_filter` restricts a call to a
chromosome or a region of a larger GDS.

## Normalize chip data before conversion (optional)

Flexible matching finds the right FAVOR row without touching the genotype file. It does not
change the file's own allele labels, which stay in the probe's order and strand. Another tool
that reads those labels as genome `REF` and `ALT`, for example when you export the data or merge
them with external summary statistics, will read them wrongly. If that matters to you, consider
normalizing the PLINK files before converting them to GDS:

- PLINK 2 `--ref-from-fa` sets `REF` from a reference FASTA at every variant where one allele
  matches it, and `--normalize` left-aligns indels.
- PLINK 2 `--ref-allele` (or PLINK 1.9 `--a2-allele`) sets the reference allele from a list.
- PLINK 1.9 `--flip` complements listed variants.

This is a recommendation, not a requirement. Normalization cannot resolve a palindromic pair
(A/T or C/G) either, since it reads the same on both strands. Whether or not you normalize, the
outcome and the rsID check are recorded, so the annotation step documents what it did.

## Appendix A. How matching works

Flexible matching uses three steps for every SNV:

1. Find a candidate FAVOR row by trying the input key and the allowed rewritings of it.
2. Compare the input's rsID with the candidate row's rsID.
3. Apply `rsid_policy` to decide whether the candidate's annotation is kept.

Indels match by the exact key under both methods, because an indel's `REF` is a stretch of
genome sequence and its swapped key names a different variant.

### Step 1: the allele lookup and its tiers

FAVOR keys every row by `CHR-POS-REF-ALT` in the VCF convention: `REF` is the base the GRCh38
reference genome carries at the position, read on the forward strand, and `ALT` is the other
base. For an SNV up to three rows exist at a position, one per alternative base. Take an input
SNV whose file labels the alleles `a1` (as `REF`) and `a2` (as `ALT`), and let `r` be the
reference base that every FAVOR row at that position shares. Write `c(x)` for the complement of
base `x`. The annotator forms the key in the four forms below, in this order, and takes the
first one that names a row. `match_tier` records which form it was.

| Tier | Key looked up | Applies when | Lookup transformation |
| --- | --- | --- | --- |
| `exact` | `a1-a2` | `a1 = r` | none |
| `swapped` | `a2-a1` | `a2 = r` | the two alleles swapped |
| `flipped` | `c(a1)-c(a2)` | `c(a1) = r` | both alleles complemented |
| `flipped_swapped` | `c(a2)-c(a1)` | `c(a2) = r` | both complemented and swapped |

Under `match_method = "exact"` only the first form is tried. A PLINK file converted with
`SeqArray::seqBED2GDS()` has its second allele in the GDS `REF` position and its first in `ALT`.
The conversion does not establish whether either is the genome's reference base.

A tier is a record of the lookup, not a verified strand. It rests on one assumption: that the
input is the reference-based substitution at that position, in some allele order and on some
strand. Under that assumption, and when FAVOR carries the row, the lookup is unique, because for
a pair of bases that is not its own complement the pair and its complement together cover A, C, G
and T, so exactly one of the four forms carries `r`. The alleles alone leave two ambiguities:

- A palindromic pair (A/T or C/G) reads the same on both strands. When it contains `r` the tier
  is `exact` or `swapped`, and that tier does not say which strand the probe used. A palindromic
  pair that lacks `r` cannot be matched.
- A pair of two non-reference alleles, whether genuine or the product of a genotyping or build
  error, is read as a strand flip, because the lookup cannot tell it apart from one.

The rsID check of step 2 supplies the evidence that the alleles cannot.

When no form names a row, or the key names no allele pair, `match_tier` carries the reason:

| Outcome | Meaning |
| --- | --- |
| `uncovered` | FAVOR has no row at the position, or a position-only key has no SNV row to aggregate |
| `unmatched_alt` | the position is covered and the reference base is found by the input's `REF` (for an indel, its first base) or, under flexible matching, by one of the rewritings, but FAVOR has no row for that allele pair |
| `unmatched_ref` | the position is covered, the input's `REF` is not `r`, and no rewriting allowed by the method finds `r` (under exact matching, no rewriting is allowed) |
| `position_only` | the key carries no alleles (`CHR-POS-NA-NA`), as Appendix B explains |
| `rsid_conflict` | a `swapped`, `flipped` or `flipped_swapped` candidate withheld by the policy (step 3) |
| `no_database` | the database has no data for the chromosome |
| `unsupported_allele` | the key has an allele other than A, C, G or T, or another unsupported form (a malformed position stops the call with an error instead) |

### Step 2: the rsID check

A FAVOR row is one substitution, and its `rsid` field holds dbSNP's identifier for that
substitution when dbSNP lists it. A matched row may carry no rsID. When it does, an rsID names
both the site and the allele pair, so agreement with the input's identifier confirms the row
whatever form found it.

The check uses the same rule for every input. For a GDS, the identifier is the node
`annotation/id`, and where it came from depends on how the GDS was made:

| Input | Where the identifier comes from |
| --- | --- |
| a GDS | `annotation/id` |
| PLINK converted with `SeqArray::seqBED2GDS()` | column 2 of the `.bim`, copied to `annotation/id` |
| VCF converted with `SeqArray::seqVCF2GDS()` | the VCF `ID` column, copied to `annotation/id` |
| a data.frame | its `rsID` column |

A genotyping chip normally carries the rsID of the intended site there, though some probes carry
vendor identifiers. A sequencing VCF often carries a missing value or a `chr:pos:ref:alt` token
unless it was annotated with dbSNP. An identifier counts as an rsID only when it is `rs`
followed by digits, in any case and ignoring surrounding whitespace. Anything else counts as no
rsID. For every matched row `rsid_check` records one value:

| `rsid_check` | Meaning |
| --- | --- |
| `same` | the input's rsID equals the matched FAVOR row's |
| `differs` | both carry an rsID and they differ |
| `chip_none` | the input carries no rsID (whether or not the row carries one) |
| `favor_none` | the input carries an rsID and the matched row carries none |
| missing | no row was matched, or the key is position-only |

The rsID is evidence about the row the alleles found. It is never used to find a row.

### Step 3: the policy

`rsid_policy` says what to do with the evidence:

- `"require"`, the default, withholds a `swapped`, `flipped` or `flipped_swapped` candidate
  whose check is `differs` or `favor_none`. The outcome becomes `rsid_conflict`, the variant
  carries no annotation, `rsid_check` keeps the value that withheld it, and `favor_key` still
  names the withheld row so that you can inspect it. An `exact` match is never withheld, since
  it used the key as given. A match checked `chip_none` is kept, since the input offers no
  evidence either way.
- `"record"` keeps every match and only records the check. Use it to see what the lookup alone
  finds, not for an annotation you will analyze.

Under the default, then, a rewritten key is accepted only when the rsID confirms it or when the
input has no rsID to check against. A sequencing data set whose identifiers are not rsIDs gets
`chip_none` on every matched variant, and nothing is withheld.

## Appendix B. Keys without alleles

A key without alleles, `CHR-POS-NA-NA`, names no row. Literature-curated variant tables often
come this way. The annotator reports such a key as `position_only` and offers two ways to fill
its annotation from the SNV rows at the position (indel rows are never included):

- `na_allele_method = "average"`, the default, gives each numeric feature the mean of its
  non-missing values over the SNV rows at the position, and leaves it missing when every value
  is missing. A string feature cannot be averaged and takes the first non-empty value.
- `"first"` takes every feature from the first SNV row at the position.

A position that carries no SNV row is `uncovered`.

Treat this as a fallback, not as annotation of the variant. The mean over three substitutions is
a quantity no variant has, and the first string value at a position is an arbitrary pick. We
recommend resolving the alleles first. When the table carries rsIDs, resolve them to alleles,
for example through dbSNP or the Ensembl REST variation endpoint, and annotate the specific
variants with exact matching. When it carries only positions, look the position up in dbSNP
before falling back to the average.

## Appendix C. A worked example

The database below is made up for illustration. It has the layout of a FAVOR v1 chunk, with a
`rsid` column on the row of the substitution dbSNP lists and empty on the other rows at the
position. It is written to a temporary directory together with its split table.

```{r build-db}
library(GLOWr)
db_dir <- tempfile("favor_demo_"); dir.create(db_dir)

favor_rows <- data.frame(
  chromosome = 22L,
  position   = c(1000L, 1000L, 1000L, 2000L, 2000L, 2000L, 2000L),
  ref_vcf    = c("G", "G", "G", "A", "A", "A", "A"),
  alt_vcf    = c("A", "C", "T", "C", "G", "T", "AT"),
  cadd_phred = c(7.25, 6.875, 7.0, 4.125, 4.25, 4.375, 9.5),
  genecode_comprehensive_category = c("exonic", "exonic", "exonic",
                                      "intronic", "intronic", "intronic", "intronic"),
  rsid       = c("rs1000", "", "", "rs2001", "", "", "rs2007"),
  stringsAsFactors = FALSE)
favor_rows$variant_vcf <- with(favor_rows, paste(chromosome, position, ref_vcf, alt_vcf, sep = "-"))

data.table::fwrite(favor_rows, file.path(db_dir, "chr22_1.csv"))
data.table::fwrite(data.frame(Chr = 22L, File_No = 1L, Start_Pos = 1L, End_Pos = 1e7),
                   file.path(db_dir, "FAVORdatabase_chrsplit.csv"))
```

The input imitates chip data: the same site in the four forms a probe can produce, a flipped pair
whose rsID differs from FAVOR's, a palindromic pair, a position-only key and an indel. The
`rsID` column is what a GDS input supplies through `annotation/id`.

```{r inputs}
chip <- data.frame(
  VarInfo = c("22-1000-G-A", "22-1000-A-G", "22-1000-C-T", "22-1000-T-C",
              "22-1000-C-T", "22-1000-C-G", "22-2000-NA-NA", "22-2000-A-AT"),
  rsID    = c("rs1000", "rs1000", "rs1000", "rs1000",
              "rs9999", "rs1000", NA, "rs2007"),
  note    = c("as FAVOR keys it", "alleles swapped", "other strand", "other strand, swapped",
              "other strand, rsID differs", "palindromic pair", "position only", "indel"),
  stringsAsFactors = FALSE)
```

Under exact matching, the first SNV and the indel are found and the position-only key is
aggregated. Under flexible matching every SNV finds a row, and the tier and the rsID check say
how. This first flexible call uses `rsid_policy = "record"` so that the table shows what the
lookup alone finds.

```{r exact-vs-flexible}
feats <- c("cadd_phred", "genecode_comprehensive_category")
exact <- annotate_favor(chip, db_dir, features = feats, match_method = "exact",
                        use_xsv = FALSE, verbose = 0)
flex  <- annotate_favor(chip, db_dir, features = feats, match_method = "flexible",
                        rsid_policy = "record", use_xsv = FALSE, verbose = 0)
cbind(chip[, c("VarInfo", "note")],
      exact = exact$match_tier,
      flexible = flex$match_tier, key = flex$favor_key, rsid_check = flex$rsid_check,
      cadd = flex$cadd_phred)
```

Three points to read off this table. First, the three rewritten forms of the first site all
carry the value of the row `22-1000-G-A`, never a mean over the position. Second, the fifth
input, the same alleles with a different rsID, is matched the same way and the check says
`differs`. Third, the position-only key averages the three SNV rows at position 2000 (`4.25`)
and leaves the indel row out.

Under the default policy two inputs are withheld: the fifth, because its rsID differs from the
row's, and the sixth, the palindromic pair, because the row it found carries no rsID to confirm
the swap. Their keys stay in `favor_key`, their annotation is missing, and nothing else changes.

```{r require}
req <- annotate_favor(chip, db_dir, features = feats, match_method = "flexible",
                      use_xsv = FALSE, verbose = 0)
cbind(chip[, "VarInfo", drop = FALSE], tier = req$match_tier, key = req$favor_key,
      rsid_check = req$rsid_check, cadd = req$cadd_phred)
```

The same content in the FAVOR 2.0 layout gives the same outcomes. This chunk runs only when
`arrow` is installed. It writes the rows as one Parquet file with the nested fields the
annotator reads (`main.cadd.phred`, `gencode.region_type`, `dbsnp.rsid`), and checks that the
outcome, the rsID check and the CADD value agree with the CSV run.

```{r parquet, eval = has_arrow}
pq_dir <- tempfile("favor2_demo_"); dir.create(pq_dir)
cadd <- data.frame(row = seq_len(nrow(favor_rows)))
cadd$cadd <- data.frame(phred = favor_rows$cadd_phred); cadd$row <- NULL
tb <- arrow::arrow_table(
  position = favor_rows$position, ref_vcf = favor_rows$ref_vcf, alt_vcf = favor_rows$alt_vcf,
  main    = arrow::StructArray$create(cadd),
  gencode = arrow::StructArray$create(
    data.frame(region_type = favor_rows$genecode_comprehensive_category)),
  dbsnp   = arrow::StructArray$create(data.frame(rsid = favor_rows$rsid)))
arrow::write_parquet(tb, file.path(pq_dir, "chromosome_22.parquet"))

pq <- annotate_favor(chip, pq_dir, features = feats, match_method = "flexible", verbose = 0)
identical(pq$match_tier, req$match_tier) && identical(pq$rsid_check, req$rsid_check) &&
  identical(pq$cadd_phred, req$cadd_phred)
```

## References

- Zhou, H., Arapoglou, T., Li, X., et al. (2023). FAVOR: functional annotation of variants
  online resource and annotator for variation across the human genome. *Nucleic Acids
  Research*, 51(D1), D1300-D1311. doi:10.1093/nar/gkac966.
- Zhou, H., Verma, V., Li, X., et al. (2026). FAVOR 2.0: A reengineered functional annotation of
  variants online resource for interpreting genomic variation. *Nucleic Acids Research*, 54(D1),
  D1405-D1414. doi:10.1093/nar/gkaf1217.
- PLINK 1.9 and PLINK 2.0 data management documentation,
  https://www.cog-genomics.org/plink/1.9/data and https://www.cog-genomics.org/plink/2.0/data.
